---
title: "AFFIP session data"
author: "Nico Bast"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cerulean
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#required packages
pkgs <- c("ggplot2", #visualization
         "kableExtra", # tables
         "data.table", # rbindlist
         "cluster", # clustering
         "pbapply", # progres bar for apply functions
         "DiagrammeR",# flowchart
         "lme4",#chain analysis
         "lmerTest", #chjain analysis
         "brms", #chain analysis - Bayesian
         "dplyr", #mutate
         "tidygraph", #network visualization
         "ggraph", #network visualization
         "igraph", #network visualization
         "tidytext"
         )

# check if required packages are installed
installed_packages = pkgs %in% rownames(installed.packages())

# install packages if not installed
if (any(installed_packages == FALSE)) {
  install.packages(pkgs[!installed_packages])
}

#loads required packages and gives error if not found
lapply(pkgs, function(pkg) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    message(paste("Package", pkg, "not found."))
  }
})


#set theme
theme_set(theme_bw())

#working directory
root_path<-"C:/Users/nico/Nextcloud/KGU/A-FFIP"


```

analysis of the session data in the [A-FFIP randomized-controlled trial](https://doi.org/10.1111/jcpp.14162). Therapist filled a form on each therapy session for each child. This included information on the A-FFIP intervention [therapy aims](#therapy-aims) (developmental area, aims, expert level) that were worked on in that session. The therapist also rated [aim outcomes](#outcome-variables) (help needed, aim mastery). The current analysis investigates whether therapy aims cluster across participants and which determinants define aim outcomes. This is achieved by concatenating sessions into [therapy aim chains](#aim-chains) based on linear mixed models, hierarchical clustering, and Bayesian Oridnal Mixed models. Further an investigation of [transitions](#chain-transitions) in therapy aims 

# summary 

**1.	Bildet sich der theoretische Ansatz des A-FFIP-Manuals mit den einzelne Förderbereichen sowie den Übungen von leicht bis Experte auch in der praktischen Arbeit mit den einzelnen Kindern ab? D.h. lernen die Kinder wie angenommen durch die Therapie aufeinander aufbauend über die entsprechenden Übungen?**

  - siehe [chain analysis](#chain-analysis) 
  - Ja, längeres Arbeiten an spezifischen Zielen (SD = 2.5 Sitzungen):
    - reduziert das [Ausmaß der Hilfestellung](#linear-mixed-help): 0.05 [0.02, 0.07] 
    - steigert die Wahrscheinlichkeit für das [Meistern der Zielebene](#linear-mixed-achieve): 0.06 [0.03, 0.09]
    - aber längere Ketten sind auch mit einem niedrigeren [Expertenlevel](#linear-mixed-mastery) verbunden, i.e. es werden im mittel dann eher leichtere Übungen durchgeführt: -0.05 [-0.08, -0.02] 
  - jeweils kontrolliert für die signifikante Kovariaten Alter, IQ, gender, Vor/Nach Corona Lockdown
  - repliziert, wenn daten beschränkt auf drei häufigsten Entwicklungsbereiche: [chain analysis restricted](#chain-analysis-restricted)
  - repliziert für [Bayesian Ordinal Mixed Models](#Bayesian-ordinal) - da outcome ratings (0,1,2) eher ordinalskaliert sind
    - 2.5 increase in chain length is associated with a 9% increase in odds of being in less help category
    - 2.5 increase in chain length is associated with about a 18% increase in odds of being in higher achieve category

**2. Ist die Anzahl der Sitzungen bis mit Therapiezielen begonnen wird von demographischen Variablen abhängig?**
  
  - Es braucht eine geringe Anzahl von Sitzungen bis mit Therapiezielen begonnen wird [ist beeinflusst](#moderating-factors-first-aims) wenn: 
    - weibliches statt männliches Geschlecht: Chi2 = 5.75, p=.01 (w: 3.73 (1.9), m: 7.98 (8.98))
    - höheres Alter der Kinder: rho = -0.31, p=.01
  - kein Einfluss von 
    - Beginn der Sitzungen vor dem Corona Lockdown 
    - Migrationsstatus der Eltern
    - Bildungsstand der Eltern
    - IQ

**3.	Mit welchen Förderbereichen wird begonnen? - Clustering**

  - siehe [Clustering](#clustering-aims) 
  - es gibt [5 Cluster](#cluster-first-aims-outcome) von Zielen mit denen begonnen wird.
    - n=21: eye contact + action planning
    - n=13: eye contact + attentional control
    - n=11: eye contact + expressive language
    - n=12: expressive langugae extended + imitation
    - n=4: attention control + gaze following
  - siehe [moderating factors](#moderating-factors-clusters) - Cluster sind nicht abhängig von demographischen Variablen 
  
**4.	Wie bauen die Förderbereiche aufeinander auf, von welchem Förderbereich wird zum anderen gewechselt? Gibt es hier spezifische Muster, sind diese vergleichbar zwischen den Kindern oder zeigen verschiedene Gruppen von Kindern in sich ähnliche aber zwischen den Gruppen unterschiedliche Muster? Gibt es hier Zusammenhänge mit Alter, nonverbaler und verbaler Entwicklung sowie autistischer Kernsymptomatik und anderen Verhaltensweisen?**

  - Diskussionspunkt: bisher liegt Fokus auf Ketten von Zielen, diese Untersuchung würde Fokus auf Transitionen von Zielen erweitern


## research aims 

- clustering of therapy aims 
- progress of patients for aims


## modelling details

- Ketten von zielen herausarbeiten
	- Mindestens zweimal bearbeitete Ziele berücksichtigen
- Outcome Ziel: 
	- Zielebene gemeistert (2)
	- Ausmaß Hilfestellung (2)
	- Zielanzahl (?)
	- Wie lange Zielkette? (predictor: chain_length)
- Pro Ziel und Kind geschaut, wie Lange an Ziel gearbeitet und mastery
- Anhand der Ketten und Qualität der ketten Clustern

## progress log 

### January 2026

1. Welche Faktoren beeinflussen Beginn der Arbeit an Zielen? - see [moderating factors](#moderating-factors)
	- Einfluss Alter, Geschlecht, Parent education, Migrationsstatus, Corona, IQ, SES
	- -> age, IQ, and lockdown haben einen Einfluss
2. Cluster der Therpapieziele zum Einstieg
	- 5 clusters that retain a language cluster
	- covariates and demographics have no influence on starting cluster
3. Session Verlauf - anhand der Kettenmodellierung
	a. Ketten auf Ebene der Ziele (REMAINS THE CURRENT FOCUS)
	b. Ketten auf Ebene der Entwicklungsbereiche  (developmental area)
	c. Ketten auf Ebene der Einzelziele (Expertlevel)

### March 2026 - descriptive analysis

1. Anzahl der Therapiesitzungen bis an Therapiezielen überhaupt gearbeitet werden kann
  - mean = 7.21 sessions, SD = 8.31 session, min-max = 2-60
2. Zeit bis an Therapiezielen überhaupt gearbeitet werden kann
  - mean = 32 days, SD = 40.3 days, min-max = 1-253

### May 2026

1. added sample description after final sample definition, see [descriptive table](#descriptive-table)
2. excluded participants with less than 10 sessions (n=2)
3. improved markdown visualization
4. removed chains length < 2
5. chain analysis restricted to three most common dev areas --> [chain analysis restricted](#chain-analysis-restricted)
6. alternative models to LMM explored --> [Bayesian Ordinal Mixed Models](#Bayesian-ordinal)
  
## TODO

  - model transitions of aims between chains
  - discuss: exclude low occurances aims (focus on prevalent aims)
  - LMM for last session in chain
  - transform mastery for Bayesian analysis (done?)
  - Markov chains for chain transitions?
  - expert level analysis for chains?
  	


# raw data

- loads the data
- based on IMBI data set (release: July 2023)

## data load

```{r load data}

#session data --> multiple rows per participant per session
project_folder<-"C:/Users/nico/Nextcloud/data_AFFIP/rct_data"
session_path<-paste0(project_folder,'/export_session_20230728.csv')
df_session<-read.csv(session_path)

#therapy data --> one row per participant per session
therapy_path<-paste0(project_folder,'/export_therapy_20230728.csv') #therapy session info - per session
df_therapy<-read.csv(therapy_path)

#gender and age in months
screen_path<-paste0(project_folder,'/export_screen_20230728.csv')
df_screen<-read.csv(screen_path) #master merge file
df_screen<-df_screen[,names(df_screen) %in% c('centno','scrnno','gender','agedemo','age_y','age_m')]

#IQ
wppsi_path<-paste0(project_folder,'/export_wppsi_20240916.csv')
df_wppsi<-read.csv(wppsi_path)
bayley_path<-paste0(project_folder,'/export_bayley_20240916.csv')
df_bayley<-read.csv(bayley_path)

#migration info
rct_path<-paste0(project_folder,'/export_rct_20230728.csv') 
df_rct<-read.csv(rct_path)

#socioeconomic status
ses_path<-paste0(project_folder,'/export_ses_20230728.csv') #socioeconomic status
df_ses<-read.csv(ses_path)

#ados symptoms
ados_path<-paste0(project_folder,'/export_ados_20230728.csv') #ADOS
df_ados<-read.csv(ados_path)


```

## IQ definition

- based on A-FFIP IQ definitions 2024 by Prof. Freitag 
- IQ is available as either Bayley or WPPSI

```{r define IQ}

#based on A-FFIP data preprocessing script

#select relevant data
df_wppsi <- df_wppsi[,names(df_wppsi) %in% c('centno', 'scrnno', 't', 'wppsi_ver','wppsidate',
                                             'vtiq', 'htiq', 'vgiq', 'asiq', 'gsiq', 'ta_wppsi_mean',
                                             'develop_age_w','verbal_iqdq_w','nonverbal_iqdq_w',
                                             'wppsi_ver','wppsidate_d','wppsidate_m','wppsidate_y')]

df_bayley<-df_bayley[,names(df_bayley) %in% c('centno','scrnno','t','baydate','develop_age',
                                              'verbal_iqdq','nonverbal_iqdq',
                                              'cognea','rezea','exprea')]

#merge data
df_iq<-merge(df_wppsi,df_bayley,by=c('centno','scrnno','t'),all.x=T,all.y=T)

#merge chronological age
df_age<-df_screen[,names(df_screen) %in% c('centno','scrnno','agedemo')]
names(df_age)[3]<-'chronage'
df_iq<-merge(df_iq,df_age,by=c('centno','scrnno'))

#Testalter (historisch)
#IQ estimate as test age /developmental age * 100
#df_iq$test_age<-with(df_iq,ifelse(is.na(ta_wppsi_mean),develop_age,ta_wppsi_mean))
df_iq$test_age<-with(df_iq,ifelse(is.na(develop_age_w),develop_age,develop_age_w))

#langugae development - not aggregated
df_iq$lang_dev<-df_iq$develop_age
df_iq$lang_dev_w<-df_iq$chronage / 100 * df_iq$asiq

#non-verbal IQ
nonverbal_iq<-df_iq$cognea / df_iq$chronage * 100
nonverbal_iq_w<-df_iq$htiq
df_iq$nonverbal_iq<-ifelse(is.na(nonverbal_iq),nonverbal_iq_w,nonverbal_iq)

#verbal IQ 
verbal_iq<-((df_iq$exprea + df_iq$rezea) / 2) / df_iq$chronage * 100
verbal_iq_w<-df_iq$vtiq
df_iq$verbal_iq<-ifelse(is.na(verbal_iq),verbal_iq_w,verbal_iq)

#full IQ 
full_iq<-((df_iq$exprea + df_iq$rezea + df_iq$cognea) / 3) / df_iq$chronage * 100
full_iq_w<-df_iq$gsiq
df_iq$full_iq<-ifelse(is.na(full_iq),full_iq_w,full_iq)

#full IQ - if nonverbal was measured with WPPSI_III and verbal with Bayley-III
# --> recovers 12 full scale IQs (!)
boolean_mixed<-!is.na(verbal_iq) & !is.na(nonverbal_iq_w)
df_iq$full_iq[boolean_mixed]<-
  ((df_iq$exprea[boolean_mixed]/df_iq$chronage[boolean_mixed] * 100) +
  (df_iq$rezea[boolean_mixed]/df_iq$chronage[boolean_mixed] * 100) +
  df_iq$htiq[boolean_mixed]) / 3
  
#developmental age
develop_age<-(df_iq$exprea + df_iq$rezea + df_iq$cognea) / 3
develop_age_w<-df_iq$chronage / 100 * df_iq$gsiq
df_iq$develop_age<-develop_age
df_iq$develop_age_w<-develop_age_w

```


## session properties

- different data frames for:
  - per session properties: df_therapy
  - per aim properties (multiple per session): df_session
- session were with caregiver (thcg),
- coding is derived from: "A-FFIP Studie Case Report Form - Therapiedokumentation"
- analysis should be limited to: 
  - Bezugsperson anwesend = 0 (ohne Bezugsperson, variable: thcg)
  - Schwerpunkt = 2 (kindzentriert, variable: focus)
  - Setting = 1 (Einzeltherapie, variable: set)
  - Diagnostiktermin = 0 (kein Diagnostiktermin, variable: diag)

```{r variable description}

#Bezugsperson anwesen
table(df_therapy$thcg) %>%
  kbl(caption = "thcg: Bezugsperson anwesend (0 = no, 1=yes)") %>%
  kable_classic(full_width = T, html_font = "Cambria") 
# --> inclusion only with 0

#Schwerpunkt
table(df_therapy$focus) %>%
  kbl(caption = "focus: Therapieschwerpunkt (1 = Elternarbeit, 2 = Kind, 3 = Austausch / Umfeld)") %>%
  kable_classic(full_width = T, html_font = "Cambria") 
# 1 = Elternarbeit
# 2 = Kind
# 3 = Austausch / Umfeld

#Setting
table(df_therapy$set) %>%
  kbl(caption = "set: Setting (1 = Einzel, 2 = Gruppe, 3 = Gespräch, 4 = Hospitation Kindergarten") %>%
  kable_classic(full_width = T, html_font = "Cambria") 
# 1 = Einzel
# 2 = Gruppe
# 3 = Gespräch
# 4 = Hospitation Kindergarten

table(df_therapy$thcg,df_therapy$focus,df_therapy$set) %>%
  kbl(caption = "Bezugsperson x Schwerpunkt x Setting",
      col.names = c('Bezugsperson','Schwerpunkt','Setting','N')) %>%
  kable_classic(full_width = T, html_font = "Cambria") 

#Diagnostiktermin 
table(df_therapy$diag) %>%
  kbl(caption = "diag: Diagnostiktermin (0 = nein)") %>%
  kable_classic(full_width = T, html_font = "Cambria") 
# 0 = kein Diagnostiktermin

```

## aim properties

- multiple aims per session
- is a different data frame in the raw data and needs to me merged with session properties
- number of aims within the session
- further defined by (practex) whether aim was actually trained
- there is no priority or rank of aims

```{r aims_per_session}

#Ziel Nummer
table(df_session$aim)  %>%
  kbl(caption = "aim: Mindestanzhal von Zielen pro Sitzung (kummulativ)") %>%
  kable_classic(full_width = T, html_font = "Cambria") 
##-->  aim number within the session (3 = third of three aims)
## 	

#Übung durchgeführt
table(df_session$practex) %>%
  kbl(caption = "practex: wurde Übung zu Ziel durchgeführt") %>%
  kable_classic(full_width = T, html_font = "Cambria")

#if Übung durchgeführt, then: Übung variiert
table(df_session$practvar)
## --> currentlky not relevant

```

## outcome variables {#outcome-variables}

- help (0-2) and achieve (0-2) define a core outcome of the analysis
- behave (0 versus 1) could be a secondary outcome variable

```{r add_vars}

#Ausmaß Hilfestellung: 0 = ohne Hilfe nicht geschafft, 1 = teilweise selbstständig, 2 = selbstständig
table(df_session$help) %>%
  kbl(caption = "help: Ausmaß Hilfestellung (0 = Hilfe notwendig, 1 = teilweise selbstständig, 2 = selbstständig)") %>%
  kable_classic(full_width = T, html_font = "Cambria") 

#Zielebene gemeistert - 0= nein, 1 = teilweise, 2 = vollständig
table(df_session$achieve) %>%
  kbl(caption = "achieve: Zielebene gemeistert (0 = nein, 1 = teilweise, 2 = vollständig)") %>%
  kable_classic(full_width = T, html_font = "Cambria") 

#herausfordernde Verhaltensweisen gezeigt - 0 = nein, 1 = ja
table(df_session$behave) %>%
  kbl(caption = "behave: Herausfordernde Verhaltensweisen (0 = nein, 1 = ja)") %>%
  kable_classic(full_width = T, html_font = "Cambria") 

```

# data set definition

- combines session and therapy data, adding demographics, and select sessions based on criteria
  - df_session = session data = multiple per session based on number of aims
  - df_therapy = therapy data = one per session

## merge data

- merges df_session (session data, p.12ff) and df_therapy (therapy data, p1-12)
- df_session: row per aim per session (n=27579)
- df_therapy: row per session (n=4731)
- adds an ID variable based on centno + scrnno
- result: df_trial

```{r merge data}

df_trial<-merge(df_session, df_therapy, 
            by.x = c("centno", "scrnno", "sessno"), 
            by.y= c("centno", "scrnno", "thsessno"))

#create IDs variable
df_trial$id<-paste(df_trial$centno,df_trial$scrnno,sep='.')

```

## add demographics

- IQ 
- gender 
- age 
- migration info 
- parental education
- autism symptoms / ADOS

```{r adding demographic data}

#adds agedemo, gender
df_trial<-merge(df_trial,df_screen, by = c("centno", "scrnno"))

#adds iq
df_iq<-df_iq[df_iq$t==1,names(df_iq)!='t'] #at earliest timepoint
df_trial<-merge(df_trial,df_iq, by = c("centno", "scrnno"))

#migration info
df_rct<-df_rct[,!(names(df_rct) %in% c('bdate_m','bdate_y'))]
df_rct<-df_rct[,names(df_rct)!='t']
df_trial<-merge(df_trial,df_rct, by = c("centno", "scrnno"), all.x = T)

#parental education and SES
selected_vars<-c("centno","scrnno","ses_edumean_t1","maxschool_t1")
df_ses<-df_ses[,names(df_ses) %in% selected_vars]
df_trial<-merge(df_trial,df_ses, by = c("centno", "scrnno"))

#adutism symptoms / ados
selected_vars<-c("centno","scrnno",'t',"adosmodule","ados_compare",'ados_sa','ados_rrb')
df_ados<-df_ados[,names(df_ados) %in% selected_vars]
df_ados<-df_ados[df_ados$t=='1',]
df_ados<-df_ados[,names(df_ados)!='t']
df_trial<-merge(df_trial,df_ados, by = c("centno", "scrnno"))

```

## included sessions

- retain only data for session that:
  - without caregiver (df_trial$thcg==0)
  - focused on child (df_trial$focus==2)
  - Einzeltherapie (df_trial$set==1)
  - no diagnostic assessment (df_trial$diag==0)
- retain only data for aims that:
  - actually trained aims (df_trail$practex==1)

```{r define_sample}

# total_aims<-nrow(df_trial)
# print(paste0('total aims in raw data: ',total_aims))
# 
# print('aims without caregiver (thcg==0):')
# table(df_trial$thcg==0)
# 
# print('aims focused on child (focus==2):')
# table(df_trial$focus==2)
# 
# print('aims in single child setting (set==1):')
# table(df_trial$set==1)
# 
# print('aims outside diagnosic sessions (diag==0):')
# table(df_trial$diag==0)
# 
# print('aims actually practiced (practex==1):')
# table(df_trial$practex==1)
# 
# #selection step
# df_trial<-df_trial[df_trial$thcg==0 & df_trial$focus==2 & df_trial$set==1 & df_trial$diag==0 & df_trial$practex==1,]
# 
# print(paste0('retained data percentage: ',round(nrow(df_trial)/total_aims,2)))
# print(paste0('retained aims: ',round(nrow(df_trial))))

total_aims <- nrow(df_trial)

# Step 1: diag == 0
keep1 <- !is.na(df_trial$diag) & df_trial$diag == 0
step1_excluded <- sum(!keep1)
df_step1 <- df_trial[keep1, ]
step1_remaining <- nrow(df_step1)

# Step 2: focus == 2
keep2 <- !is.na(df_step1$focus) & df_step1$focus == 2
step2_excluded <- sum(!keep2)
df_step2 <- df_step1[keep2, ]
step2_remaining <- nrow(df_step2)

# Step 3: thcg == 0
keep3 <- !is.na(df_step2$thcg) & df_step2$thcg == 0
step3_excluded <- sum(!keep3)
df_step3 <- df_step2[keep3, ]
step3_remaining <- nrow(df_step3)

# Step 4: set == 1
keep4 <- !is.na(df_step3$set) & df_step3$set == 1
step4_excluded <- sum(!keep4)
df_step4 <- df_step3[keep4, ]
step4_remaining <- nrow(df_step4)

# Step 5: practex == 1
keep5 <- !is.na(df_step4$practex) & df_step4$practex == 1
step5_excluded <- sum(!keep5)
df_step5 <- df_step4[keep5, ]
step5_remaining <- nrow(df_step5)

# Final selected dataset
df_trial_selected <- df_step5

print(paste0("Total aims in raw data: ", total_aims))
print(paste0("Retained data percentage: ", round(nrow(df_trial_selected) / total_aims, 2)))
print(paste0("Retained aims: ", nrow(df_trial_selected)))

#flowchart
grViz(paste0("
digraph flowchart {
  graph [layout = dot, rankdir = TB]
  node [shape = box, style = rounded, fontname = Helvetica]
  
  main0 [label = 'Raw data\nN = ", total_aims, "', style=filled, fillcolor=LightBlue]
  main1 [label = 'diag == 0\nTherapy sessions\nN = ", step1_remaining, "', style=filled, fillcolor=LightBlue]
  main2 [label = 'focus == 2\nChild-focused\nN = ", step2_remaining, "', style=filled, fillcolor=LightBlue]
  main3 [label = 'thcg == 0\nNo caregiver present\nN = ", step3_remaining, "', style=filled, fillcolor=LightBlue]
  main4 [label = 'set == 1\nSingle-child setting\nN = ", step4_remaining, "', style=filled, fillcolor=LightBlue]
  main5 [label = 'practex == 1\nActually practiced aims\nFinal N = ", step5_remaining, "', style=filled, fillcolor=Green]

  ex1 [label = 'Excluded\ndiag != 0 or NA\nn = ", step1_excluded, "', style=filled, fillcolor=MistyRose]
  ex2 [label = 'Excluded\nfocus != 2 or NA\nn = ", step2_excluded, "', style=filled, fillcolor=MistyRose]
  ex3 [label = 'Excluded\nthcg != 0 or NA\nn = ", step3_excluded, "', style=filled, fillcolor=MistyRose]
  ex4 [label = 'Excluded\nset != 1 or NA\nn = ", step4_excluded, "', style=filled, fillcolor=MistyRose]
  ex5 [label = 'Excluded\npractex != 1 or NA\nn = ", step5_excluded, "', style=filled, fillcolor=MistyRose]

  main0 -> main1
  main1 -> main2
  main2 -> main3
  main3 -> main4
  main4 -> main5

  main0 -> ex1 [dir=none]
  main1 -> ex2 [dir=none]
  main2 -> ex3 [dir=none]
  main3 -> ex4 [dir=none]
  main4 -> ex5 [dir=none]
}
"))

#retain data
df_trial<-df_trial_selected

```

## pandemic break

- first shutdown in Germany: 22/03/2020 - 04/05/2020
- calculate a corona state based on lockdown onset (after_lockdown=T == session after 22/03/20)

```{r pandemic_break}

  #lockdown as officially defined:
  lockdown_start_germany<-as.Date("22/03/20", format = "%d/%m/%y")
  lockdown_end_germany<-as.Date("04/05/20", format = "%d/%m/%y")

  #across all
  df_trial$sdate_<-as.Date(df_trial$sdate_, format = "%d/%m/%y")
  ggplot(df_trial,aes(sdate_,sessno,color=id))+geom_point()+
    geom_vline(xintercept=lockdown_start_germany,color='red')+
    geom_vline(xintercept=lockdown_end_germany,color='red')+
    theme(legend.position = "none")+
    labs(x='year',y='session number',title='sessions over duration of the study',subtitle='each color represents a participant')
    
hist(df_trial$sdate_,150,main='sessions across time of study')

#define variable
df_trial$after_lockdown<-ifelse(df_trial$sdate_>lockdown_start_germany,T,F)
  
print(paste0('proportion of therapy session after first lockdown (22/03/20): ',
             round(nrow(df_trial[df_trial$after_lockdown==T,])/nrow(df_trial),2)))
  
```

# sample description

see [sample description](#descriptive-table). Excludes particitpants with less than 10 sessions (n=2). Participants have different [number of sessions](#session-numbers)

## exclude participants

- exclude participants with less than 10 session 
  - only one session (id = 1.32)
  - only 6 sessions (id = 4.23)

```{r exclude_participants}

df_trial<-df_trial[!df_trial$id=='1.32',]
df_trial<-df_trial[!df_trial$id=='4.23',]


```

## participants & sessions overview {#session-numbers}

- shows participants ids (centno.scrnno)
- sessions by participant
- define an ID variable for df_trial

```{r number_of_participants}

#participants
print('number of participants:')
length(table(interaction(df_trial$centno,df_trial$scrnno,drop=T)))

#session by participant
sessions_by_participant<-as.numeric(
  by(df_trial$sessno,interaction(df_trial$centno,df_trial$scrnno,drop=T),function(x){length(unique(x))}))
hist(sessions_by_participant,main = 'sessions by participant')

participant_IDs<-names(table(interaction(df_trial$centno,df_trial$scrnno,drop=T)))

df_ids<-data.frame(participant_IDs,sessions_by_participant)
df_ids<-df_ids[order(df_ids$sessions_by_participant),]

df_ids %>%
  kbl(caption = "Sessions by participant") %>%
  kable_classic(full_width = T, html_font = "Cambria") 

```

## sample description table {#descriptive-table}

```{r sample_description_table}

description_variables<-c('centno','scrnno','agedemo','gender',
                         'nonverbal_iq','verbal_iq','full_iq',
                         'migration','language','ses_edumean_t1','maxschool_t1')

#create data frame with unique ids
df_sample<-df_trial[!duplicated(interaction(df_trial$centno,df_trial$scrnno,drop=T)),names(df_trial) %in% description_variables]

#table for categorical variables
gender<-table(df_sample$gender)
migration<-table(df_sample$migration)
language<-table(df_sample$language)

df_sample_categorical<-data.frame(rbind(gender,migration,language))
row_names<-c('gender (0 = male, 1=female)','migration history of parents (0=yes, 1=no)',
             'language spoken at home (0=German, 1=other)')
df_sample_categorical<-cbind(row_names,df_sample_categorical)
names(df_sample_categorical)<-c('variable','0','1')

df_sample_categorical %>%
  kbl(caption = "Table 1A. sample description for categorical variables",row.names = F) %>%
  kable_classic(full_width = T, html_font = "Cambria") 


#table for numeric variables
df_sample_numeric<-rbind(
  psych::describe(df_ids$sessions_by_participant), #note: utilizes different data set
  psych::describe(df_sample$agedemo),
  psych::describe(df_sample$nonverbal_iq),
  psych::describe(df_sample$verbal_iq),
  psych::describe(df_sample$full_iq),
  psych::describe(df_sample$ses_edumean_t1),
  psych::describe(df_sample$maxschool_t1)
)

df_sample_numeric<-df_sample_numeric[,c('n','mean','sd','min','max')]
df_sample_numeric<-round(df_sample_numeric,2)

row.names(df_sample_numeric)<-c('number of sessions','age (in m)','nonverbal IQ','verbal IQ','full IQ',
                                'parents education (mean)','parent education (max)')

df_sample_numeric %>%
  kbl(caption = "Table 1B. Sample description for numeric variables") %>%
  kable_classic(full_width = F, html_font = "Cambria") %>%
  footnote(general = c("parent education: 0 = no school (ISCED 0)","1 = school not completed (ISCED 1)","2 = completed special school (ISCED 1)","3 = Hauptschulabschluss (ISCED 2A)",'4 = Realschulabschluss/Berufsfachschule (ISCED3B)',"5 = Hochschulreife/Abitur (ISCED 3A)"))

```

# therapy aims overview {#therapy-aims}

- can be more than one per session
- encoded by "aimno" variable (e.g. 10.6.2.1)
  - first or two numbers define developmental area (e.g. 9, 10.2-10.6)
  - first three numbers define specific therapy aim (e.g. 10.6.2) --> defines the [therapy aim chains](#aim-chains)
  - last number defines expert level (1,2,3) --> rather as outcome variable
  
## developmental area

- Entwicklungsbereiche (9.2 - 10.6)
- can be more than one per session

```{r developmental area}

## Ziele overall
#table(df_trial$aimno)

#create data
dev_area<-substr(df_trial$aimno,1,8)
dev_area<-ifelse(nchar(dev_area)==5,substr(dev_area,1,1),dev_area)
dev_area<-ifelse(nchar(dev_area)==8,substr(dev_area,1,4),dev_area)
dev_area<-ifelse(grepl("A",dev_area),'A',dev_area)
dev_area<-ifelse(grepl("B",dev_area),'B',dev_area)
dev_area<-ifelse(grepl("-8",dev_area),NA,dev_area)

dev_area_char<-dev_area
dev_area<-as.factor(dev_area)
levels(dev_area)<-c('communication & speech (10.2)','interaction & play (10.3)','emotion (10.4)','cognitive ability (10.5)','daily living skills (10.6)','basic abilities (9)','introduction (A)','reduction of unwanted behaviors (B)')

df_trial$dev_area<-dev_area
df_trial$dev_area_char<-dev_area_char

#visualize data
dev_area_table<-data.frame(sort(table(df_trial$dev_area),decreasing = TRUE))
dev_area_table_char<-data.frame(sort(table(df_trial$dev_area_char),decreasing = TRUE))
dev_area_table$dev_area_char<-dev_area_table_char$Var1
dev_area_table<-dev_area_table[,c(1,3,2)]

dev_area_table %>%
  kbl(caption = "developmental areas across sessions") %>%
  kable_classic(full_width = T, html_font = "Cambria") 

ggplot(df_trial,aes(x=dev_area,fill=dev_area))+geom_bar()+theme(axis.text.x=element_blank())

```

## therapy aims

- an welchen zielen wurde gearbeitet?
- can be more than one per session

```{r therapy aims}

## Ziele overall
#table(df_trial$aimno)

aims<-substr(df_trial$aimno,1,8)
aims<-ifelse(nchar(aims)==8,substr(aims,1,6),aims)
aims<-ifelse(nchar(aims)==5,substr(aims,1,3),aims)
aims<-ifelse(grepl("-8",aims),NA,aims)

aims_rank<-aims
aims_rank<-ifelse(grepl('10',aims_rank),substr(aims_rank,6,6),aims_rank)
aims_rank<-ifelse(grepl('9',aims_rank),substr(aims_rank,3,3),aims_rank)
aims_rank<-ifelse(grepl('A',aims_rank),NA,aims_rank)
aims_rank<-ifelse(grepl('B',aims_rank),NA,aims_rank)
df_trial$aims_rank<-aims_rank

aims_char<-aims
aims<-as.factor(aims)
levels(aims)<-c('eye contact','passive vocabulary','gesture','expressive language (supported)',
                'expressive language extended','questions & answers','making contact',
                
                'shared attention','exchange','sharing','non-interactive play','simple shared play',
                'rule-based play','symbolic play','role play',
                
                'emotion recognition','emotion expression','emotion regulation',
                
                'classification','characteristics/opposites','memory',
                'drawing','quantitative understanding','theory of mind',
                'common knowledge',
                
                'daily routines','toilet training',
                'problem solving','awareness of danger','spatial awareness','temporal orientation',
                
                'attention control','gaze following',
                'imitation','representation','action planing','self awareness',
                
                'testing reinforcement','building therapeutic alliance','situational control','separation from caregiver',
                
                'aggression','stereotypic behavior','alternatives','anxiety','avoidance'
                
                )

df_trial$aims<-aims
df_trial$aims_char<-aims_char

aims_table<-data.frame(sort(table(df_trial$aims),decreasing = TRUE))
aims_table_char<-data.frame(sort(table(df_trial$aims_char),decreasing = TRUE))
aims_table$aims_char<-aims_table_char$Var1
aims_table<-aims_table[,c(1,3,2)]

  aims_table %>%
  kbl(caption = "occurance of aims across sessions") %>%
  kable_classic(full_width = F, html_font = "Cambria") 

#commented out as aim rank is not relevant  
  
# aims_table_byrank<-as.matrix(table(df_trial$dev_area,df_trial$aims_rank))
# 
#   aims_table_byrank %>%
#   kbl(caption = "aim ranks by developmental area across sessions") %>%
#   kable_classic(full_width = F, html_font = "Cambria") 

```

# first therapy aims 

- extract first session per participant and according aims
- remove introduction sessions
- the number of aims in the first session can be more than one per child 

```{r create data of first session therapy aims}

#create data
df_firstaims<-df_trial[df_trial$dev_area!='introduction (A)',]
df_firstaims<-split(df_firstaims,df_firstaims$id)
df_firstaims<-lapply(df_firstaims,function(x){
  earliest_session<-min(x$sessno)
  x<-x[x$sessno==earliest_session,]
  return(x)
})
df_firstaims<-rbindlist(df_firstaims)

```

## first aims overview

- top 3: eye contact (10.2.1), imitation (9.4), gaze following (9.3)

```{r first aims across number of aims}

#table overview
aims_table<-sort(table(droplevels(df_firstaims$aims)),decreasing = TRUE)
aims_table<-data.frame(aims_table)

aims_char_table<-sort(table(df_firstaims$aims_char),decreasing = TRUE)
aims_table<-data.frame(names(aims_char_table),aims_table)
names(aims_table)<-c('code','aim','occurances')

  aims_table %>%
  kbl(caption = "aims in the first session after introdcution") %>%
  kable_classic(full_width = F, html_font = "Cambria") 

#aims by participant
table_aimspart<-by(df_firstaims$aims,df_firstaims$id,function(x){table(droplevels(x))})

table_aimspart<-sapply(table_aimspart,dimnames)
max_length<-max(sapply(table_aimspart,length))
table_aimspart<-lapply(table_aimspart,function(x){
  current_length<-length(x)
  new_vector<-rep(NA,max_length)
  new_vector[1:current_length]<-x
  return(new_vector)})

table_aimspart<-rbind.data.frame(table_aimspart)

table_aimspart <- data.frame(t(table_aimspart))
names(table_aimspart) <- paste('goal', seq_along(table_aimspart))
row.names(table_aimspart) <- substr(row.names(table_aimspart), 2, nchar(row.names(table_aimspart)))

print('discussion point: Factor in the number of aims per participant?')

table_aimspart[order(table_aimspart[,1]),] %>%
  kbl(caption = "aims in the first session after introdcution by participant") %>%
  kable_classic(full_width = F, html_font = "Cambria") 

```

## time till first aims

```{r, time span to first session}

#extract date of first session
df_firstsession<-df_therapy[df_therapy$thsessno==1,]
df_firstsession<-df_firstsession[,names(df_firstsession) %in% c('centno','scrnno','thdate_')]
names(df_firstsession)<-c('centno','scrnno','date_first_session')

#add - adds df_firstaims that already removed A/introduction codings
df_firstaims<-merge(df_firstaims,df_firstsession,by =c("centno", "scrnno"))

#convert to date
df_firstaims$sdate_<- as.Date(df_firstaims$sdate_, format = "%d/%m/%y")
df_firstaims$date_first_session<- as.Date(df_firstaims$date_first_session, format = "%d/%m/%y")

#select only one row per participant
df_firstaims_uniqueID<-df_firstaims[!duplicated(df_firstaims$id),]

time_to_first_session<-data.frame(df_firstaims_uniqueID$sdate_-df_firstaims_uniqueID$date_first_session)
names(time_to_first_session)<-'time_in_d'

# table(time_to_first_session) %>%
#    kbl(caption = "time in days from introduction to session with first aims") %>%
#   kable_classic(full_width = T, html_font = "Cambria") 
# 
# ggplot(time_to_first_session,aes(x=time_in_d))+geom_bar()+xlab('time in days to first session with aims')


df_date_first_session<-data.frame(df_firstaims_uniqueID$id,time_to_first_session$time_in_d,
                                  df_firstaims_uniqueID$sessno,df_firstaims_uniqueID$sdate_)

df_date_first_session$days_by_session<-round(df_date_first_session$time_to_first_session.time_in_d/
                                               df_date_first_session$df_firstaims_uniqueID.sessno,2)

names(df_date_first_session)<-c('id','days to first session','session number','date of first session','days / session')
df_date_first_session<-df_date_first_session[order(df_date_first_session$'days to first session'),]

#descriptive data
summary(df_date_first_session$`days to first session`)
sd(df_date_first_session$`days to first session`)

#table
df_date_first_session %>%
   kbl(caption = "first session with aims: date and time by IDs", row.names = FALSE) %>%
   kable_classic(full_width = T, html_font = "Cambria")


```

## sessions till first aims

- Berechnung der Anzahl der Therapiesitzungen, die Kind gebraucht hat, bis an Therapiezielen überhaupt gearbeitet werden kann - Kriterium: Anzahl der „A“-Sitzungen zu Beginn der Therapie“

```{r first session as therapy session}

psych::describe(df_firstaims$sessno[!duplicated(df_firstaims$id)])

#session number after introduction   
hist(df_firstaims$sessno[!duplicated(df_firstaims$id)],
     main = 'session number when first aims',20)  

table(df_firstaims$sessno[!duplicated(df_firstaims$id)]) %>%
  kbl(caption = "Session number after introduction (after 'A')") %>%
  kable_classic(full_width = T, html_font = "Cambria") 

```

### outlier investigation

- id 2.4 has 60 sessions till first aims
- caregiver was present during first 60 sessions
- has 30 sessions without caregiver after that - leave in analysis

```{r outlier_analysis first session}

outlier_sessions<-df_session[df_session$centno=='2' & df_session$scrnno=='4',]
outlier_sessions$sdate_<-as.Date(outlier_sessions$sdate_, format = "%d/%m/%y")
ggplot(outlier_sessions,aes(sdate_,sessno,color=aimno))+geom_point()+labs(title= 'outlier participant concerning duration of introduction',subtitle = 'color encodes different aims that were worked on for this participant while caregiver present')

outlier_sessions<-outlier_sessions[order(outlier_sessions$sessno),]

outlier_sessions2<-df_therapy[df_therapy$centno=='2' & df_therapy$scrnno=='4',]
table(outlier_sessions2$thcg)

```

## moderating factors {#moderating-factors-first-aims}

- which factors influence the start of working on the first therapy aims
- consider age, gender, parental education, migration status, corona, IQ, SES

- for non-parametric testing: age, IQ, and lockdown have an effect on session number
  - age: spearman correlation: rho = -0.31, p = .014
  - IQ: spearman correlation: rho = -0.26, p = .045
  - lockdown: session before lockdown: m/SD = 7.03/10.45, session after lockdown: m/SD = 7.57/5.18, p = .47

--> consider those as covariates in testing?

```{r moderating factors on first aim}

#select only one row per participant
df_firstaims_uniqueID<-df_firstaims[!duplicated(df_firstaims$id),]

#nonparametric testing - categorical variables
print('gender:')
with(df_firstaims_uniqueID,kruskal.test(sessno ~ as.factor(gender)))
with(df_firstaims_uniqueID,by(sessno,gender,psych::describe))
print('migration status:')
with(df_firstaims_uniqueID,kruskal.test(sessno ~ as.factor(migration), data = df_firstaims_uniqueID))
print('parental education:')
with(df_firstaims_uniqueID,kruskal.test(sessno ~ as.factor(maxschool_t1), data = df_firstaims_uniqueID))
print('lockdown status:')
with(df_firstaims_uniqueID,kruskal.test(sessno ~ as.factor(after_lockdown), data = df_firstaims_uniqueID))

#--> significant influence of lockdown - sessions after lockdown take longer to reach aims
with(df_firstaims_uniqueID,by(sessno,after_lockdown,psych::describe))

#nonparametric testing - continous variables
print('age:')
with(df_firstaims_uniqueID, cor.test(agedemo, sessno, method = "spearman",exact=F))
print('IQ:')
with(df_firstaims_uniqueID,cor.test(nonverbal_iq, sessno, method = "spearman",exact=F))
print('SES:')
with(df_firstaims_uniqueID,cor.test(ses_edumean_t1, sessno, method = "spearman",exact=F))

```

# clustering first aims {#clustering-aims}

- Cluster method: Gower Distance Hierarchical Clustering
- Gower Distance = useful when comapring a mix of categorical and numeric variables
- consider the first 7 therapy aims of every participant at the first session
- considers missing data (not all participants have equal number of aims)
- categorical data type

```{r first aim clustering, fig.width=10, fig.height=6}

# Cut into clusters (e.g., 3 clusters)
number_of_clusters<-3
print(paste('assumed number of clusters:',number_of_clusters))

#translate - 7 first aims by participant
table_aimspart_factor<-lapply(table_aimspart,as.factor)
table_aimspart_factor<-data.frame(table_aimspart_factor[1],
                                  table_aimspart_factor[2],
                                  table_aimspart_factor[3],
                                  table_aimspart_factor[4],
                                  table_aimspart_factor[5],
                                  table_aimspart_factor[6],
                                  table_aimspart_factor[7])

rownames(table_aimspart_factor)<-row.names(table_aimspart)


# Compute Gower distance
gower_dist <- daisy(table_aimspart_factor, metric = "gower")
print('clustering metric - gower distance between pairwise comparisons:')
table(round(gower_dist,2))

# Perform hierarchical clustering
hc <- hclust(gower_dist, method = "ward.D2")
clusters <- cutree(hc, k = number_of_clusters)

# Add cluster labels to your data
table_aimspart_factor$start_cluster <- clusters

print('5 clusters combine language aims better into distinct cluster')

# table_aimspart_factor[order(table_aimspart_factor[,8]),] %>%
#   kbl(caption = "aims in the first session and starting CLUSTERS after introduction by participant") %>%
#   kable_classic(full_width = F, html_font = "Cambria") 

# Labels for dendrogram
dendro_labels <- paste0(rownames(table_aimspart_factor), " [Cluster ", clusters, "]")

# visualize cluster
plot(hc, labels = dendro_labels, main = "Dendrogram of Hierarchical Clustering")
rect.hclust(hc, k = number_of_clusters, border = "red")  # Draw boxes for clusters

table(table_aimspart_factor$start_cluster)

#add clusters to data
id<-row.names(table_aimspart_factor)
start_cluster<-table_aimspart_factor$start_cluster
df_cluster<-data.frame(id,start_cluster)
df_firstaims<-merge(df_firstaims,df_cluster,by='id')
df_firstaims$start_cluster<-as.factor(df_firstaims$start_cluster)

df_firstaims_uniqueID<-data.frame(df_firstaims[!duplicated(df_firstaims$id),])
df_firstaims_uniqueID<-df_firstaims_uniqueID[,names(df_firstaims_uniqueID) %in% c('id','start_cluster')]
df_trial<-merge(df_trial,df_firstaims_uniqueID,by='id')

```

## describe clusters {#cluster-first-aims-outcome}

- 5 most common aims in different clusters for first therapy aims
- labels the clusters 

```{r describe initial clusters}

#show 5 most common aims in different clusters
cluster_aims<-by(table_aimspart_factor[,1:7],table_aimspart_factor$start_cluster,
   function(x){sort(table(droplevels(unlist(x))),decreasing=T)[1:5]})
cluster_aims<-lapply(cluster_aims,data.frame)
cluster_aims<-do.call(cbind,cluster_aims)
names(cluster_aims)<-rep(c('label','frequency'),number_of_clusters)

participants_in_clusters<-as.numeric(table(table_aimspart_factor$start_cluster))
header_label<-rep(2,number_of_clusters)
names(header_label)<-paste('cluster',1:number_of_clusters,'(n=',participants_in_clusters,')')

cluster_aims %>%
  kbl(caption = "5 most common aims in the first session by clusters") %>%
  kable_classic(full_width = F, html_font = "Cambria") %>%
  add_header_above(header_label)


```

## moderating clusters {#moderating-factors-clusters}

- non-parametric Fisher's exact Test (categorical outcome = cluster, categorical predictor)
- no moderator had an influence

```{r moderating clusters}

#select only one row per participant
df_firstaims_uniqueID<-df_firstaims[!duplicated(df_firstaims$id),]
table(df_firstaims_uniqueID$start_cluster)

#nonparametric testing - categorical variables
print('gender:')
with(df_firstaims_uniqueID,table(start_cluster, gender))
with(df_firstaims_uniqueID,fisher.test(table(start_cluster, gender)))
print('migration status:')
with(df_firstaims_uniqueID,table(start_cluster, migration))
with(df_firstaims_uniqueID,fisher.test(table(start_cluster, migration)))
print('parental education:')
with(df_firstaims_uniqueID,table(start_cluster, maxschool_t1))
with(df_firstaims_uniqueID,fisher.test(table(start_cluster, maxschool_t1)))
print('lockdown status:')
with(df_firstaims_uniqueID,table(start_cluster, after_lockdown))
with(df_firstaims_uniqueID,fisher.test(table(start_cluster, after_lockdown)))

#--> significant influence of lockdown - sessions after lockdown take longer to reach aims
with(df_firstaims_uniqueID,by(sessno,after_lockdown,psych::describe))

#nonparametric testing - continous variables
print('age:')
with(df_firstaims_uniqueID, kruskal.test(agedemo ~ start_cluster))
print('IQ:')
with(df_firstaims_uniqueID, kruskal.test(full_iq ~ start_cluster))
print('SES:')
with(df_firstaims_uniqueID, kruskal.test(ses_edumean_t1 ~ start_cluster))

```


# therapy aims expert level

- expert level defines a type of mastery of a specific aim
- last number defines expert level (1,2,3)
- can be utilized as outcome measure of training 

```{r therapy aims expert level}

expert_level<-substr(df_trial$aimno,5,8)
expert_level<-ifelse(nchar(expert_level)==4,substr(expert_level,4,4),expert_level)
expert_level<-ifelse(nchar(expert_level)==0,NA,expert_level)

expert_level<-as.factor(expert_level)
levels(expert_level)<-c('beginner','intermediate','expert')

df_trial$expert_level<-expert_level

with(df_trial,table(dev_area,expert_level))

ggplot(df_trial,aes(x=sessno,fill=expert_level))+geom_bar()+facet_wrap(~dev_area)


```

# chains of therapy aims {#aim-chains}

- identify chains of consecutive working on specific therapy aims 
- at least 2 consecutive sessions with one aim should be later considered

## chain functions

- code below define functions to extract chains of aims
- use "show" to examine code

```{r chain functions}

#function to extract outcome data for chains of a single aim within one participant
#outcome data includes: achieve, help
fun_chain_metrics<-function(single_chain){
#chain static info
single_chain_outcome<-single_chain[1,names(single_chain) %in% c('id','centno','scrnno','aims','aims_char',
                                                                'dev_area','dev_area_char','start_cluster',
                                                                'gender','agedemo','develop_age','verbal_iq',
                                                                'nonverbal_iq','full_iq','test_age',
                                                                'migration','maxschool_t1','ses_edumean_t1',
                                                                'after_lockdown')]

#chain characterization
chain_length<-nrow(single_chain)
chain_first_session<-min(single_chain$sessno)
chain_last_session<-max(single_chain$sessno)
chain_first_date<-min(single_chain$sdate_)
chain_last_date<-max(single_chain$sdate_)
chain_after_lockdown_percentage<-mean(single_chain$after_lockdown)
chain_behave_mean<-mean(single_chain$behave)

#chain outcome metrics
chain_achieve_mean<-mean(single_chain$achieve)
chain_achieve_first<-single_chain$achieve[single_chain$sessno==chain_first_session]
chain_achieve_last<-single_chain$achieve[single_chain$sessno==chain_last_session]
chain_achieve_0<-sum(single_chain$achieve==0)
chain_achieve_1<-sum(single_chain$achieve==1)
chain_achieve_2<-sum(single_chain$achieve==2)
chain_help_mean<-mean(single_chain$help)
chain_help_first<-single_chain$help[single_chain$sessno==chain_first_session]
chain_help_last<-single_chain$help[single_chain$sessno==chain_last_session]
chain_help_0<-sum(single_chain$help==0)
chain_help_1<-sum(single_chain$help==1)
chain_help_2<-sum(single_chain$help==2)
chain_expertlevel_beginner<-sum(single_chain$expert_level=='beginner')
chain_expertlevel_intermediate<-sum(single_chain$expert_level=='intermediate')
chain_expertlevel_expert<-sum(single_chain$expert_level=='expert')
chain_expertlevel_last<-single_chain$expert_level[single_chain$sessno==chain_last_session]

single_chain_variables<-data.frame(
  #chain characterization
  chain_length,
  chain_first_session,
  chain_last_session,
  chain_first_date,
  chain_last_date,
  chain_after_lockdown_percentage,
  chain_behave_mean,
  #chain outcome metrics
  chain_achieve_mean,
  chain_achieve_first,
  chain_achieve_last,
  chain_achieve_0,
  chain_achieve_1,
  chain_achieve_2,
  chain_help_mean,
  chain_help_first,
  chain_help_last,
  chain_help_0,
  chain_help_1,
  chain_help_2,
  chain_expertlevel_beginner,
  chain_expertlevel_intermediate,
  chain_expertlevel_expert,
  chain_expertlevel_last
  )

single_chain_outcome<-data.frame(single_chain_outcome,single_chain_variables)
return(single_chain_outcome)

}

#function that loops fun_chain_metrics across aims by participant
fun_chain_loop_ids_aims<-function(single_participant){
  
#testing:  
#single_participant<-list_trial[[22]]
  
#identify aims
all_aims<-names(table(droplevels(single_participant$aims)))

#identifies chains for participant - loops across all aims of a participant
single_participant_chains<-list()
for(i in seq_along(all_aims)){
single_participant_single_aim<-single_participant[single_participant$aims==all_aims[i],] 
# Identify breaks
group_id <- cumsum(c(TRUE, diff(single_participant_single_aim$sessno) != 1))
# Split into list of chains
chains <- split(single_participant_single_aim, group_id)
#apply fun_chain_metrics to each list element
chains_of_single_aim<-lapply(chains,fun_chain_metrics)
#return to data.frame
chains_of_single_aim<-data.table::rbindlist(chains_of_single_aim)
#retrieve number of chains for single aim
chains_of_single_aim$aim_chain_nr <- seq_len(nrow(chains_of_single_aim))
single_participant_chains[[i]]<-chains_of_single_aim
}

names(single_participant_chains)<-all_aims #names list of chains by aims
single_participant_chains<-data.table::rbindlist(single_participant_chains) #turn list of chains to data.frame  
return(single_participant_chains)

}

```

## chain data extraction

- splits data by participants
- splits single participant by aims
- create chains for every participant (for loop)
- create mastery variable based on expert level
- remove chains with length < 2 (removes about 50% of chains; i.e. half of chains are 1 session chains)

```{r prepare data of chains}

#split by participant
list_trial<-split(df_trial,df_trial$id)
list_trial<-lapply(list_trial,function(x){x[order(x$sessno),]})

# applies fucntions above in loop (participants) of loops (aims per participant)
list_chains<-pblapply(list_trial,fun_chain_loop_ids_aims)
df_chains<-rbindlist(list_chains)

#create additional variable 
df_chains$chain_mastery_mean<-with(df_chains,
             ((chain_expertlevel_beginner*0)+(chain_expertlevel_intermediate*1)+(chain_expertlevel_expert*2))/
               (chain_expertlevel_beginner+chain_expertlevel_intermediate+chain_expertlevel_expert))

#remove chain length < 2 (removes 50% of data)
df_chains<-df_chains[df_chains$chain_length>1,]

```

## chain description

```{r chain description}

table(df_chains$chain_length) %>%
  kbl(caption = "total chain length") %>%
  kable_classic(full_width = F, html_font = "Cambria")

hist(with(df_chains,by(chain_length,aims,mean)),main = 'mean chain length')

table_chainlength<-data.frame(with(df_chains,by(chain_length,aims,function(x){round(mean(x),2)})))
chain_occurances<-as.numeric(with(df_chains,by(chain_length,aims,length)))
table_chainlength<-data.frame(table_chainlength,chain_occurances)

table_chainlength<-table_chainlength[order(table_chainlength[,2],decreasing = T),]
names(table_chainlength)<-c('mean chain length','chain occurances')

table_chainlength %>%
  kbl(caption = "mean chain length by aim") %>%
  kable_classic(full_width = F, html_font = "Cambria")

```

## chain visualization

```{r chain visualization}

ggplot(df_chains[!df_chains$dev_area_char %in% c('A','B')],
       aes(chain_first_session,id,
           alpha=chain_last_session-chain_first_session,
           color=dev_area))+
  geom_point()+
  facet_wrap(~dev_area)

ggplot(df_chains[!df_chains$dev_area_char %in% c('A','B')],
       aes(chain_first_session,weight=chain_last_session-chain_first_session,
           fill=dev_area))+
  geom_density(aes(y = after_stat(count)), adjust = 1)+ #weighted by occurances
  labs(x='session number',
       title = 'density of chains across sessions weighted by chain length')+
  facet_wrap(~dev_area)

```

## chain analysis {#chain-analysis}

- main predictor: chain_length
- main outcome:
  - help ("Ausmaß Hilfestellung (0-2)")
  - achieve ("Zielebene gemeistert (0-2)")
  - mastery ("aim expert level (0-2)")
- only consider chains not disrupted by lockdown; complete before or after

```{r chain analysis data inspection}

#describe outcome variables
selected_vars<-names(df_chains) %in% c('chain_length','chain_help_mean','chain_achieve_mean','chain_mastery_mean')
desc<-apply(df_chains[,..selected_vars],2,psych::describe)
desc<-do.call(rbind,desc)

desc_tbl <- desc %>%
  tibble::rownames_to_column("variable") %>%
  select(variable, n, mean, sd, median, min, max, skew, kurtosis) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

description<-c('predictor','outcome: "Zielebene gemeistert"','outcome: "Ausmaß Hilfestellung"','outcome: "Expertenlevel/einzelne Ziele"')

desc_tbl<-cbind(description,desc_tbl)

desc_tbl %>%
  kbl(caption = "descriptive statistics: core variables") %>%
  kable_classic(full_width = T, html_font = "Cambria") 

#non-normal distribution
par(mfrow = c(2, 2))

hist(df_chains$chain_length,main='chain length',xlab='chain_length')
hist(df_chains$chain_help_mean,main='Ausmaß Hilfestellung',xlab='chain_help_mean (mean of session chain)')
hist(df_chains$chain_achieve_mean,main='Zielebene gemeistert',xlab='chain_achieve_mean (mean of session chain)')
hist(df_chains$chain_mastery_mean,main='Expertenlevel / einzelne Ziele',xlab='chain_mastery_mean (mean of session chain)')

par(mfrow = c(1, 1))

```


```{r chain analysis data preparation}

#prepare data 
lmm_data<-df_chains[df_chains$chain_after_lockdown_percentage %in% c(0,1),]

# function to plot model as APA table
fun_table_lmm<-function(lmm,caption){

lmm_summary<-data.frame(round(summary(lmm)$coefficients,3))
names(lmm_summary)<-c('estimate','SE','df','t-value','p-value')
variable<-c('intercept','chain length','full IQ','gender','age','after lockdown')
lmm_summary<-data.frame(variable,lmm_summary)

lmm_summary %>%
  kableExtra::kbl(caption = caption,row.names = F) %>%
  kableExtra::kable_classic(full_width = F, html_font = "Cambria") 

}

```

### linear mixed models 

#### LMM help {#linear-mixed-help}

```{r chain analysis frequentist help mean}

#mean
lmm<-lmer(scale(chain_help_mean)~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
            as.factor(chain_after_lockdown_percentage)+(1|id),data=lmm_data)
fun_table_lmm(lmm,caption="help as mean across sessions: effect of chain length")

plot(lmm)
qqnorm(resid(lmm))
qqline(resid(lmm))

#last
lmm<-lmer(scale(chain_help_last)~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
            as.factor(chain_after_lockdown_percentage)+(1|id),data=lmm_data)
fun_table_lmm(lmm,caption="help in last session of chain: effect of chain length")

plot(lmm)
qqnorm(resid(lmm))
qqline(resid(lmm))


```

#### LMM achieve {#linear-mixed-achieve}

```{r chain analysis frequentist achieve}

#mean
lmm<-lmer(scale(chain_achieve_mean)~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
            as.factor(chain_after_lockdown_percentage)+(1|id),data=lmm_data)
fun_table_lmm(lmm,caption="achieve as mean across sessions: effect of chain length")

plot(lmm)
qqnorm(resid(lmm))
qqline(resid(lmm))

#last
lmm<-lmer(scale(chain_achieve_last)~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
            as.factor(chain_after_lockdown_percentage)+(1|id),data=lmm_data)
fun_table_lmm(lmm,caption="achieve in last session of chain: effect of chain length")

plot(lmm)
qqnorm(resid(lmm))
qqline(resid(lmm))

```

#### LMMM mastery {#linear-mixed-mastery}

```{r chain analysis frequentist mastery}

lmm<-lmer(scale(chain_mastery_mean)~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
            as.factor(chain_after_lockdown_percentage)+(1|id),data=lmm_data)
fun_table_lmm(lmm,caption="mastery outcome: effect of chain length")

plot(lmm)
qqnorm(resid(lmm))
qqline(resid(lmm))


lmm_data$chain_mastery_last<-as.numeric(lmm_data$chain_expertlevel_last)
table(lmm_data$chain_mastery_last,lmm_data$chain_expertlevel_last)

lmm<-lmer(scale(chain_mastery_last)~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
            as.factor(chain_after_lockdown_percentage)+(1|id),data=lmm_data)
fun_table_lmm(lmm,caption="mastery outcome: effect of chain length")

plot(lmm)
qqnorm(resid(lmm))
qqline(resid(lmm))


```

### linear mixed models - common dev areas {#chain-analysis-restricted}

- only extract the most common developmental areas
- effects are confirmed when data is restricted to three most common dev areas

```{r chain analysis frequentist in common dev areas}

lmm_data_res<-lmm_data[lmm_data$dev_area %in% c('communication & speech (10.2)',
                                                'basic abilities (9)',
                                                'interaction & play (10.3)'),]

lmm<-lmer(scale(chain_help_mean)~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
            as.factor(chain_after_lockdown_percentage)+(1|id),data=lmm_data_res)
fun_table_lmm(lmm,caption="help outcome: effect of chain length")


lmm<-lmer(scale(chain_achieve_mean)~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
            as.factor(chain_after_lockdown_percentage)+(1|id),data=lmm_data_res)
fun_table_lmm(lmm,caption="achieve outcome: effect of chain length")

lmm<-lmer(scale(chain_mastery_mean)~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
            as.factor(chain_after_lockdown_percentage)+(1|id),data=lmm_data_res)
fun_table_lmm(lmm,caption="mastery outcome: effect of chain length")

```

### Bayesian ordinal mixed model {#Bayesian-ordinal}

- assumes ordinal data
- consider the last session in a chain and its rating

```{r chain analysis Bayesian}

lmm_data$chain_help_last<-as.factor(lmm_data$chain_help_last)
lmm_data$chain_help_last<-ordered(lmm_data$chain_help_last, c(0,1,2))

# fit<-brm(chain_help_last~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
#             as.factor(chain_after_lockdown_percentage)+(1|id),
#          data=lmm_data,
#          family = cumulative('logit'),
#          chains = 4,
#          cores = 4
#          )
# 
# summary(fit)
# saveRDS(fit, file = paste0(getwd(),"/session data/data/brms_model_chain_help.rds"))


fit <- readRDS(paste0(root_path,"/session data/data/brms_model_chain_help.rds"))
summary(fit)
print('a 1 SD increase in chain length is associated with about a 9% increase in odds of being in less help category:')
exp(0.09)


# lmm_data$chain_achieve_last<-ordered(lmm_data$chain_achieve_last, c(0,1,2))
# fit2<-brm(chain_achieve_last~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
#             as.factor(chain_after_lockdown_percentage)+(1|id),
#          data=lmm_data,
#          family = cumulative('logit'),
#          chains = 4,
#          cores = 4
#          )
# 
# summary(fit2)
# saveRDS(fit2, file = paste0(getwd(),"/session data/data/brms_model_chain_achieve.rds"))

fit2 <- readRDS(paste0(root_path,"/session data/data/brms_model_chain_achieve.rds"))
summary(fit2)
print('a 1 SD increase in chain length is associated with about a 18% increase in odds of being in higher achieve category:')
exp(0.17)


####chain_mastery_last

# lmm_data$chain_expertlevel_last<-ordered(lmm_data$chain_expertlevel_last, c('beginner','intermediate','expert'))
# fit3<-brm(chain_expertlevel_last~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
#             as.factor(chain_after_lockdown_percentage)+(1|id),
#          data=lmm_data,
#          family = cumulative('logit'),
#          chains = 4,
#          cores = 4
#          )
# 
# summary(fit3)
# saveRDS(fit3, file = paste0(getwd(),"/session data/data/brms_model_chain_mastery.rds"))

fit3 <- readRDS(paste0(root_path,"/session data/data/brms_model_chain_mastery.rds"))
summary(fit3)
print('a 1 SD increase in chain length is not associated with change in odds of being in another mastery level')
exp(-0.02)

```

# transitions of aims {#chain-transitions}

- aim chains occur across the therapy, we want to extract aim transitions between chains
- retrieves the aim of the next and previous chain
- previous and next chains are defined by lower or higher therapy session number in the last session of a chain

```{r add prev next chains}
# add previous / next chain information per participant
# order chains per participant by starting session and compute lag/lead
df_chains <- df_chains %>%
  dplyr::group_by(id) %>%
  dplyr::arrange(chain_last_session, .by_group = TRUE) %>%
  dplyr::mutate(
    prev_chain_aim = dplyr::lag(as.character(aims)),
    next_chain_aim = dplyr::lead(as.character(aims)),
    prev_chain_length = dplyr::lag(chain_length),
    next_chain_length = dplyr::lead(chain_length),
    prev_chain_dev_area = dplyr::lag(as.character(dev_area_char)),
    next_chain_dev_area = dplyr::lead(as.character(dev_area_char)),
    prev_same_aim = ifelse(is.na(prev_chain_aim), NA, prev_chain_aim == as.character(aims)),
    next_same_aim = ifelse(is.na(next_chain_aim), NA, next_chain_aim == as.character(aims))
  ) %>%
  dplyr::ungroup()

```

## transitions across participants

- general transitions across all participants
- overall to many transitions, this requires filtering below

```{r chain transition matrix}
# create transition matrix: current aim -> next aim
# only include chains with non-NA next_chain_aim (i.e., exclude last chains per participant)
df_transitions <- df_chains[!is.na(df_chains$next_chain_aim), ]

# build transition matrix (count)
transition_matrix_count <- table(df_transitions$aims, df_transitions$next_chain_aim)

# order rows and columns by total occurrences (descending)
transition_matrix_count<-transition_matrix_count[order(rowSums(transition_matrix_count), 
                              decreasing = TRUE), 
                        order(colSums(transition_matrix_count), decreasing = TRUE)]

# convert to proportions (row-wise: probability of transitioning FROM each aim)
transition_matrix_prop <- prop.table(transition_matrix_count, margin = 1)

# round for display
transition_matrix_prop_display <- round(transition_matrix_prop, 2)

# display as kable
transition_matrix_prop_display %>%
  kbl(caption = "Chain transition probabilities: likelihood of transitioning FROM aim (rows) TO aim (columns) order by total occurances") %>%
  kable_classic(full_width = FALSE, html_font = "Cambria")

```

- filtered based on a minimum number of occurrences (5% of all transitions)
- to focus on more prevalent transitions

```{r transition filter}

# filter transitions based on minimum occurrences
# inclusion threshold: aims must appear in at least 5% of all transitions
minimal_number_of_occurances <- round(0.05 * nrow(df_transitions))

aim_counts <- table(c(as.character(df_transitions$aims), as.character(df_transitions$next_chain_aim)))
common_aims <- names(aim_counts[aim_counts > minimal_number_of_occurances])

df_transitions_filtered <- df_transitions[
  (as.character(df_transitions$aims) %in% common_aims) & 
  (as.character(df_transitions$next_chain_aim) %in% common_aims), 
]

cat("Filtered to", length(common_aims), "aims with >", minimal_number_of_occurances, "total occurrences\n")
cat("Remaining transitions:", nrow(df_transitions_filtered), "out of", nrow(df_transitions), "\n\n")
cat("Percentage of chains that repeat same aim:", 
    round(mean(df_transitions_filtered$next_same_aim, na.rm = TRUE) * 100, 1), "%\n")
cat("Percentage of chains that switch to different aim:", 
    round(mean(df_transitions_filtered$next_same_aim == FALSE, na.rm = TRUE) * 100, 1), "%\n")

```

- transition heatmap shows the occurances and proportion of transitions

```{r transition heatmap, fig.width=12, fig.height=12}

# rebuild count and proportion matrices from filtered data
transition_matrix_count <- table(df_transitions_filtered$aims, df_transitions_filtered$next_chain_aim)
transition_matrix_prop  <- prop.table(transition_matrix_count, margin = 1)

# prepare long format
transition_matrix_long <- as.data.frame.table(transition_matrix_count)
names(transition_matrix_long) <- c('from_aim', 'to_aim', 'count')
transition_matrix_long$proportion <- as.data.frame.table(transition_matrix_prop)[, 3]
transition_matrix_long <- transition_matrix_long[
  !is.na(transition_matrix_long$from_aim) & !is.na(transition_matrix_long$to_aim), ]

# order rows and columns by total occurrences (descending)
aim_order_overall <- names(sort(aim_counts[names(aim_counts) %in% common_aims],
                                decreasing = TRUE))
transition_matrix_long <- transition_matrix_long |>
  dplyr::mutate(
    from_aim = factor(from_aim, levels = aim_order_overall),
    to_aim   = factor(to_aim,   levels = aim_order_overall)
  )

# top 10 most prevalent transitions weighted by count * proportion
# (rewards transitions that are both frequent and conditionally probable)
transition_matrix_long |>
  dplyr::filter(as.character(from_aim) != as.character(to_aim)) |>
  dplyr::mutate(
    score = count * proportion,
    proportion = round(proportion, 3)
  ) |>
  dplyr::arrange(dplyr::desc(score)) |>
  dplyr::slice_head(n = 10) |>
  dplyr::select(
    `From Aim`   = from_aim,
    `To Aim`     = to_aim,
    `Count`      = count,
    `Proportion` = proportion,
    `Score (n × p)` = score
  ) |>
  dplyr::mutate(`Score (n × p)` = round(`Score (n × p)`, 1)) |>
  kbl(caption = "Top 10 most prevalent transitions weighted by count × row-proportion (self-transitions excluded)") |>
  kable_classic(full_width = FALSE, html_font = "Cambria")


# visualize as heatmap
ggplot(transition_matrix_long, aes(x = to_aim, y = from_aim, fill = proportion)) +
  geom_tile(color = 'white') +
  scale_fill_gradient(low = 'white', high = 'steelblue') +
  geom_text(aes(label = count), size = 3) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(title = paste('Chain Transitions (aims with >',
                     minimal_number_of_occurances,
                     'occurrences) From current aim (row) to next aim (column)',
       x = 'Next Aim', y = 'Current Aim', fill = 'transition likelihood',
       caption = 'Grey marked rows indicate that transitions to other aims were rare with less than 50 occurrences.'))

```

## chain transition network

```{r chain transition network, message=FALSE, warning=FALSE, fig.width=10, fig.height=9}

# thresholds
min_prop <- 0.1
min_occurrences <- round(0.05 * nrow(df_transitions_filtered))

frequent_aims <- names(aim_counts[aim_counts > min_occurrences])

edges <- transition_matrix_long |>
  filter(
    proportion >= min_prop,
    as.character(from_aim) != as.character(to_aim),
    as.character(from_aim) %in% frequent_aims,
    as.character(to_aim) %in% frequent_aims
  ) |>
  mutate(from = as.character(from_aim), to = as.character(to_aim)) |>
  select(from, to, weight = proportion, count)

# lookup: aim -> developmental area (using full descriptive labels)
aim_dev_lookup <- df_chains |>
  dplyr::distinct(aims, dev_area) |>
  dplyr::mutate(name = as.character(aims), dev_area = as.character(dev_area)) |>
  dplyr::select(name, dev_area)

dev_area_levels <- sort(unique(aim_dev_lookup$dev_area))
dev_area_colors <- setNames(RColorBrewer::brewer.pal(length(dev_area_levels), "Set2"),
                            dev_area_levels)

# node sizes from total occurrences (as current aim), with dev_area
node_totals <- df_transitions_filtered |>
  count(aims, name = "n_total") |>
  rename(name = aims) |>
  mutate(name = as.character(name)) |>
  left_join(aim_dev_lookup, by = "name")

graph <- as_tbl_graph(edges, directed = TRUE) |>
  left_join(node_totals, by = "name")

ggraph(graph, layout = "fr") +
  geom_edge_fan(
    aes(width = weight, alpha = weight),
    arrow = arrow(length = unit(3, "mm"), type = "closed"),
    end_cap = circle(5, "mm"),
    show.legend = TRUE
  ) +
  scale_edge_width(range = c(0.5, 3), name = "Proportion") +
  scale_edge_alpha(range = c(0.4, 0.9), guide = "none") +
  geom_node_point(aes(size = n_total, color = dev_area)) +
  geom_node_text(aes(label = n_total), size = 2.5, color = "white", fontface = "bold") +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_size_continuous(name = "Total occurrences", range = c(4, 14)) +
  scale_color_manual(values = dev_area_colors, name = "Developmental Area",
                     drop = FALSE) +
  labs(title = "Chain Transition Network",
       subtitle = paste0("Edges: proportion \u2265 ", min_prop,
                         "; aims with > ", min_occurrences, " occurrences; self-loops excluded")) +
  theme_graph()
```

## transitions between clusters

```{r cluster-prep, message=FALSE, warning=FALSE}
library(ggraph)
library(tidygraph)

min_prop_cl <- 0.10

# lookup: aim -> developmental area
aim_dev_lookup <- df_chains |>
  dplyr::distinct(aims, dev_area) |>
  dplyr::mutate(name = as.character(aims), dev_area = as.character(dev_area)) |>
  dplyr::select(name, dev_area)

# consistent color palette across all cluster plots
dev_area_levels <- sort(unique(aim_dev_lookup$dev_area))
dev_area_colors <- setNames(RColorBrewer::brewer.pal(length(dev_area_levels), "Set2"),
                            dev_area_levels)

cluster_ids <- sort(unique(as.character(df_chains$start_cluster)))

# build all per-cluster data once; reused by the three display chunks below
cluster_data <- list()

for (cl in cluster_ids) {

  df_cl    <- df_chains[as.character(df_chains$start_cluster) == cl, ]
  df_tr_cl <- df_cl[!is.na(df_cl$next_chain_aim), ]

  min_occurrences_cl <- round(0.05 * nrow(df_tr_cl))

  aim_counts_cl  <- table(c(as.character(df_tr_cl$aims),
                             as.character(df_tr_cl$next_chain_aim)))
  common_aims_cl <- names(aim_counts_cl[aim_counts_cl > min_occurrences_cl])

  df_tr_filt <- df_tr_cl[
    (as.character(df_tr_cl$aims)           %in% common_aims_cl) &
    (as.character(df_tr_cl$next_chain_aim) %in% common_aims_cl), ]

  if (nrow(df_tr_filt) == 0) {
    cluster_data[[cl]] <- NULL
    next
  }

  tm_count <- table(df_tr_filt$aims, df_tr_filt$next_chain_aim)
  tm_prop  <- prop.table(tm_count, margin = 1)

  tm_long <- as.data.frame.table(tm_count)
  names(tm_long) <- c("from_aim", "to_aim", "count")
  tm_long$proportion <- as.data.frame.table(tm_prop)[, 3]
  tm_long <- tm_long[!is.na(tm_long$from_aim) & !is.na(tm_long$to_aim), ]

  aim_order_cl <- names(sort(aim_counts[names(aim_counts) %in% common_aims_cl],
                             decreasing = TRUE))
  tm_long <- tm_long |>
    dplyr::mutate(
      from_aim = factor(from_aim, levels = aim_order_cl),
      to_aim   = factor(to_aim,   levels = aim_order_cl)
    )

  edges_cl <- tm_long |>
    dplyr::filter(
      proportion >= min_prop_cl,
      as.character(from_aim) != as.character(to_aim)
    ) |>
    dplyr::mutate(from = as.character(from_aim), to = as.character(to_aim)) |>
    dplyr::select(from, to, weight = proportion, count)

  node_totals_cl <- df_tr_filt |>
    dplyr::count(aims, name = "n_total") |>
    dplyr::rename(name = aims) |>
    dplyr::mutate(name = as.character(name)) |>
    dplyr::left_join(aim_dev_lookup, by = "name")

  graph_cl <- if (nrow(edges_cl) > 0) {
    as_tbl_graph(edges_cl, directed = TRUE) |>
      dplyr::left_join(node_totals_cl, by = "name")
  } else NULL

  cluster_data[[cl]] <- list(
    tm_long            = tm_long,
    edges_cl           = edges_cl,
    graph_cl           = graph_cl,
    node_totals_cl     = node_totals_cl,
    min_occurrences_cl = min_occurrences_cl,
    common_aims_cl     = common_aims_cl,
    n_transitions      = nrow(df_tr_cl)
  )
}
```

### Top 10 transitions per cluster

```{r cluster-tables, results='asis'}
cat('<div style="display:flex; flex-wrap:wrap; gap:1.5em; align-items:flex-start;">')

for (cl in cluster_ids) {
  d <- cluster_data[[cl]]
  cat('<div>')

  if (is.null(d)) {
    cat(paste0("<p><strong>Cluster ", cl, ":</strong> insufficient data</p>"))
  } else {
    tbl <- d$tm_long |>
      dplyr::filter(as.character(from_aim) != as.character(to_aim)) |>
      dplyr::mutate(score = count * proportion, proportion = round(proportion, 3)) |>
      dplyr::arrange(dplyr::desc(score)) |>
      dplyr::slice_head(n = 10) |>
      dplyr::select(
        `From` = from_aim,
        `To`   = to_aim,
        `n`    = count,
        `p`    = proportion,
        `n×p`  = score
      ) |>
      dplyr::mutate(`n×p` = round(`n×p`, 1)) |>
      kbl(format = "html", caption = paste0("Cluster ", cl)) |>
      kable_classic(full_width = FALSE, html_font = "Cambria")
    print(tbl)
  }

  cat('</div>')
}

cat('</div>')
```

### Transition heatmaps per cluster

::: {style="overflow-x: auto; white-space: nowrap; padding-bottom: 8px;"}

```{r cluster-heatmaps, message=FALSE, warning=FALSE, fig.width=18, fig.height=6, out.width="4200px", dpi=100}
heat_plots <- lapply(cluster_ids, function(cl) {
  d <- cluster_data[[cl]]
  if (is.null(d)) return(ggplot() + labs(title = paste0("Cluster ", cl, ": no data")))

  ggplot(d$tm_long, aes(x = to_aim, y = from_aim, fill = proportion)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "steelblue") +
    geom_text(aes(label = count), size = 2.5) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          axis.text   = element_text(size = 8)) +
    labs(
      title   = paste0("Cluster ", cl),
      x       = "Next Aim", y = "Current Aim", fill = "Proportion",
      caption = paste0("Threshold: >", d$min_occurrences_cl, " occurrences (5% of n=",
                       d$n_transitions, ")")
    )
})

patchwork::wrap_plots(heat_plots, ncol = number_of_clusters) +
  patchwork::plot_layout(guides = "collect") +
  patchwork::plot_annotation(title = "Chain Transition Heatmaps by Cluster")
```

:::

### Transition networks per cluster

```{r cluster-networks, message=FALSE, warning=FALSE, fig.width=14, fig.height=30}

net_plots <- lapply(cluster_ids, function(cl) {
  d <- cluster_data[[cl]]
  if (is.null(d) || is.null(d$graph_cl)) {
    return(ggplot() + labs(title = paste0("Cluster ", cl, ": no data")) + theme_void())
  }

  ggraph(d$graph_cl, layout = "fr") +
    geom_edge_fan(
      aes(width = weight, alpha = weight),
      arrow    = arrow(length = unit(3, "mm"), type = "closed"),
      end_cap  = circle(5, "mm"),
      show.legend = TRUE
    ) +
    scale_edge_width(range = c(0.5, 3), name = "Proportion") +
    scale_edge_alpha(range = c(0.4, 0.9), guide = "none") +
    geom_node_point(aes(size = n_total, color = dev_area)) +
    geom_node_text(aes(label = n_total), size = 2.5, color = "white", fontface = "bold") +
    geom_node_text(aes(label = name), repel = TRUE, size = 3) +
    scale_size_continuous(name = "Total occurrences", range = c(4, 14)) +
    scale_color_manual(values = dev_area_colors, name = "Developmental Area",
                       drop = FALSE) +
    labs(title = paste0("Cluster ", cl)) +
    theme_graph()
})

patchwork::wrap_plots(net_plots, nrow = number_of_clusters) +
  patchwork::plot_layout(guides = "collect") +
  patchwork::plot_annotation(
    title    = "Chain Transition Networks by Cluster",
    subtitle = paste0("Edges: proportion \u2265 ", min_prop_cl, "; self-loops excluded")
  )

```


# transition network comparison 

```{r centrality-comparison, message=FALSE, warning=FALSE, fig.width=12, fig.height=7}

# ── extract centrality metrics from pre-built cluster_data ─────────────────
centrality_list <- list()
net_stats_list  <- list()

for (cl in cluster_ids) {

  d <- cluster_data[[cl]]
  if (is.null(d) || is.null(d$graph_cl)) next

  g_c <- d$graph_cl

  # node-level centrality metrics
  centrality_list[[cl]] <- g_c |>
    dplyr::mutate(
      # weighted in/out-strength: sum of proportions on incoming/outgoing edges
      in_strength  = centrality_degree(mode = "in",  weights = weight),
      out_strength = centrality_degree(mode = "out", weights = weight),
      # unweighted betweenness: structural bridge role
      betweenness  = centrality_betweenness(directed = TRUE),
      # pagerank with proportion weights: overall flow-weighted importance
      pagerank     = centrality_pagerank(weights = weight)
    ) |>
    as_tibble() |>
    dplyr::mutate(cluster = paste0("Cluster ", cl))

  # network-level summary stats
  ig_c <- igraph::as.igraph(g_c)
  net_stats_list[[cl]] <- tibble::tibble(
    Cluster             = paste0("Cluster ", cl),
    Nodes               = igraph::vcount(ig_c),
    Edges               = igraph::ecount(ig_c),
    Density             = round(igraph::edge_density(ig_c), 3),
    `Mean betweenness`  = round(mean(igraph::betweenness(ig_c, directed = TRUE)), 2),
    `Mean PageRank`     = round(mean(igraph::page_rank(ig_c,
                                       weights = igraph::E(ig_c)$weight)$vector), 4)
  )
}

all_centrality <- dplyr::bind_rows(centrality_list)
net_stats      <- dplyr::bind_rows(net_stats_list)
```

```{r centrality-net-table}
# ── network-level summary table ────────────────────────────────────────────
net_stats |>
  kbl(caption = "Network-level properties per cluster") |>
  kable_classic(full_width = FALSE, html_font = "Cambria")
```

```{r centrality-pagerank, fig.width=18, fig.height=6}
# ── PageRank heatmap: aims × clusters ─────────────────────────────────────
# show aims that appear in at least two clusters, ranked by mean PageRank
top_aims_pr <- all_centrality |>
  dplyr::group_by(name) |>
  dplyr::summarise(mean_pr = mean(pagerank), n_clusters = dplyr::n()) |>
  dplyr::filter(n_clusters >= 2) |>
  dplyr::slice_max(mean_pr, n = 20) |>
  dplyr::pull(name)

all_centrality |>
  dplyr::filter(name %in% top_aims_pr) |>
  dplyr::mutate(name = forcats::fct_reorder(name, pagerank, mean)) |>
  ggplot(aes(x = cluster, y = name, fill = pagerank)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "steelblue", name = "PageRank") +
  facet_grid(dev_area ~ ., scales = "free_y", space = "free_y") +
  labs(
    title = "PageRank by aim and cluster",
    subtitle = "Top 20 aims present in \u22652 clusters; grouped by developmental area",
    x = NULL, y = NULL
  ) +
  theme(
    strip.text.y    = element_text(angle = 0, hjust = 0, size = 7),
    axis.text.y     = element_text(size = 8),
    panel.spacing.y = unit(2, "pt")
  )
```

```{r centrality-betweenness, fig.width=18, fig.height=6}
# ── betweenness: top 6 bridge nodes per cluster ────────────────────────────
all_centrality |>
  dplyr::group_by(cluster) |>
  dplyr::slice_max(betweenness, n = 6) |>
  dplyr::ungroup() |>
  # encode cluster into label so reorder() works within facets
  dplyr::mutate(label = reorder(paste0(cluster, "___", name), betweenness)) |>
  ggplot(aes(x = betweenness, y = label, fill = dev_area)) +
  scale_y_discrete(labels = \(x) sub(".*___", "", x)) +
  geom_col() +
  scale_fill_manual(values = dev_area_colors, name = "Developmental Area") +
  facet_wrap(~ cluster, scales = "free_y", ncol = 3) +
  labs(
    title = "Top 6 aims by betweenness centrality per cluster",
    subtitle = "Betweenness identifies aims acting as bridges between other therapy goals",
    x = "Betweenness centrality", y = NULL
  )
```

## transitions with session progression

- divide session to early (1-30), mid (31-60), late (61-90) therapy
- compare networks

```{r transitions sessions prepare}

df_chains <- df_chains %>%
mutate(session_cat = case_when(
chain_first_session <= 30 ~ "early",
chain_first_session > 30 & chain_first_session <= 60 ~ "mid",
chain_first_session > 60 ~ "late",
TRUE ~ NA_character_
))

sess_cats <- unique(as.character(df_chains$session_cat))

# build all per-cluster data once; reused by the three display chunks below
cluster_data <- list()

for (sess in sess_cats) {

  #select one cat
  df_cl    <- df_chains[as.character(df_chains$session_cat) == sess, ]
  
  #at least 5% of occurances
  min_occurrences_cl <- round(0.05 * nrow(df_cl))

  aim_counts_cl  <- table(c(as.character(df_cl$aims),
                             as.character(df_cl$next_chain_aim)))
  common_aims_cl <- names(aim_counts_cl[aim_counts_cl > min_occurrences_cl])

  df_tr_filt <- df_cl[
    (as.character(df_cl$aims)           %in% common_aims_cl) &
    (as.character(df_cl$next_chain_aim) %in% common_aims_cl), ]

  if (nrow(df_tr_filt) == 0) {
    cluster_data[[cl]] <- NULL
    next
  }

  tm_count <- table(df_tr_filt$aims, df_tr_filt$next_chain_aim)
  tm_prop  <- prop.table(tm_count, margin = 1)

  tm_long <- as.data.frame.table(tm_count)
  names(tm_long) <- c("from_aim", "to_aim", "count")
  tm_long$proportion <- as.data.frame.table(tm_prop)[, 3]
  tm_long <- tm_long[!is.na(tm_long$from_aim) & !is.na(tm_long$to_aim), ]

  aim_order_cl <- names(sort(aim_counts[names(aim_counts) %in% common_aims_cl],
                             decreasing = TRUE))
  tm_long <- tm_long |>
    dplyr::mutate(
      from_aim = factor(from_aim, levels = aim_order_cl),
      to_aim   = factor(to_aim,   levels = aim_order_cl)
    )

  edges_cl <- tm_long |>
    dplyr::filter(
      proportion >= min_prop_cl,
      as.character(from_aim) != as.character(to_aim)
    ) |>
    dplyr::mutate(from = as.character(from_aim), to = as.character(to_aim)) |>
    dplyr::select(from, to, weight = proportion, count)

  node_totals_cl <- df_tr_filt |>
    dplyr::count(aims, name = "n_total") |>
    dplyr::rename(name = aims) |>
    dplyr::mutate(name = as.character(name)) |>
    dplyr::left_join(aim_dev_lookup, by = "name")

  graph_cl <- if (nrow(edges_cl) > 0) {
    as_tbl_graph(edges_cl, directed = TRUE) |>
      dplyr::left_join(node_totals_cl, by = "name")
  } else NULL

  cluster_data[[sess]] <- list(
    tm_long            = tm_long,
    edges_cl           = edges_cl,
    graph_cl           = graph_cl,
    node_totals_cl     = node_totals_cl,
    min_occurrences_cl = min_occurrences_cl,
    common_aims_cl     = common_aims_cl,
    n_transitions      = nrow(df_tr_cl)
  )
}


```

### Top 10 transitions per session progression category

```{r cluster-tables sessions, results='asis'}
cat('<div style="display:flex; flex-wrap:wrap; gap:1.5em; align-items:flex-start;">')

for (sess in sess_cats) {
  d <- cluster_data[[sess]]
  cat('<div>')

  if (is.null(d)) {
    cat(paste0("<p><strong>therapy progression: ", sess, ":</strong> insufficient data</p>"))
  } else {
    tbl <- d$tm_long |>
      dplyr::filter(as.character(from_aim) != as.character(to_aim)) |>
      dplyr::mutate(score = count * proportion, proportion = round(proportion, 3)) |>
      dplyr::arrange(dplyr::desc(score)) |>
      dplyr::slice_head(n = 10) |>
      dplyr::select(
        `From` = from_aim,
        `To`   = to_aim,
        `n`    = count,
        `p`    = proportion,
        `n×p`  = score
      ) |>
      dplyr::mutate(`n×p` = round(`n×p`, 1)) |>
      kbl(format = "html", caption = paste0("therapy progression: ", sess)) |>
      kable_classic(full_width = FALSE, html_font = "Cambria")
    print(tbl)
  }

  cat('</div>')
}

cat('</div>')
```

### Transition networks per session progression category

```{r cluster-networks sessions, message=FALSE, warning=FALSE, fig.width=14, fig.height=30}

net_plots <- lapply(sess_cats, function(sess) {
  d <- cluster_data[[sess]]
  if (is.null(d) || is.null(d$graph_cl)) {
    return(ggplot() + labs(title = paste0("Cluster ", sess, ": no data")) + theme_void())
  }

  ggraph(d$graph_cl, layout = "fr") +
    geom_edge_fan(
      aes(width = weight, alpha = weight),
      arrow    = arrow(length = unit(3, "mm"), type = "closed"),
      end_cap  = circle(5, "mm"),
      show.legend = TRUE
    ) +
    scale_edge_width(range = c(0.5, 3), name = "Proportion") +
    scale_edge_alpha(range = c(0.4, 0.9), guide = "none") +
    geom_node_point(aes(size = n_total, color = dev_area)) +
    geom_node_text(aes(label = n_total), size = 2.5, color = "white", fontface = "bold") +
    geom_node_text(aes(label = name), repel = TRUE, size = 3) +
    scale_size_continuous(name = "Total occurrences", range = c(4, 14)) +
    scale_color_manual(values = dev_area_colors, name = "Developmental Area",
                       drop = FALSE) +
    labs(title = paste0("therapy progression category: ", sess)) +
    theme_graph()
})

patchwork::wrap_plots(net_plots, nrow = length(sess_cats)) +
  patchwork::plot_layout(guides = "collect") +
  patchwork::plot_annotation(
    title    = "Chain Transition Networks by therapy progression category",
    subtitle = paste0("Edges: proportion \u2265 ", min_prop_cl, "; self-loops excluded")
  )

```

- clustern based on clusters and quality of clusters
- Prüfung der weiteren Angaben der Therapieziel im Verlauf der Therapie -> Stabilität bzw. Wechsel in andere Kategorie.

```{r comparing clusters}

anova(lmer(chain_length~start_cluster+(1|id),data=df_chains))
anova(lmer(chain_achieve_last~start_cluster+(1|id),data=df_chains))
anova(lmer(chain_help_last~start_cluster+(1|id),data=df_chains))

anova(lmer(chain_help_last~start_cluster*chain_length+(1|id),data=df_chains))

hist(df_chains$chain_first_session)


```
