---
title: "AFFIP session data"
author: "Nico Bast"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cerulean
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#required packages
require(ggplot2) #visualization
require(kableExtra) #tables
require(data.table) #rbindlist
require(cluster) #clustering
require(pbapply) #progres bar for apply functions
library(DiagrammeR) #flowchart
require(lme4) #chain analysis
require(lmerTest) #chain analysis
require(brms) #chain analysis - Bayesian

#set theme
theme_set(theme_bw())

#working directory
root_path<-"C:/Users/nico/Nextcloud/KGU/A-FFIP"


```

# research aims 

- clustering of therapy aims 
- progress of patients for aims

## summary 

1.	Bildet sich der theoretische Ansatz des A-FFIP-Manuals mit den einzelne Förderbereichen sowie den Übungen von leicht bis Experte auch in der praktischen Arbeit mit den einzelnen Kindern ab? D.h. lernen die Kinder wie angenommen durch die Therapie aufeinander aufbauend über die entsprechenden Übungen?

  - siehe [chain analysis](#chain-analysis) 
  - Ja, längeres Arbeiten an spezifischen Zielen reduziert das Ausmaß der Hilfestellung und steigert die Wahrscheinlichkeit für das meistern der Zielebene. 
  - Aber längere Ketten sind auch mit einem niedrigeren Expertenlevel verbunden. 
  - Dies ist kontrolliert für die signifikante Kovariaten Alter, IQ, Vor/Nach Corona Lockdown
  - repliziert, wenn daten beschränkt auf drei häufigsten Entwicklungsbereiche: [chain analysis restricted](#chain-analysis-restricted)

2.	Mit welchen Förderbereichen wird begonnen? - Clustering

  - siehe [Clustering](#clustering-aims) 
  - es gibt 5 Cluster von Zielen mit denen begonnen wird.  
  - Diese sind unabhängig von Kovariaten
  
3. Gibt es [bzgl. der CLuster] Zusammenhänge mit Alter, nonverbaler und verbaler Entwicklung sowie autistischer Kernsymptomatik und anderen Verhaltensweisen?  
  
  - siehe [moderating factors](#moderating-factors) 
  - die Anzahl der Sitzungen bis mit Therapiezielen begonnen wird ist beeinflusst durch: 
  - höheres Alter der Kinder, höherer non-verbaler IQ, und Beginn der Sitzungen vor dem Corona Lockdown 

3.	Wie bauen die Förderbereiche aufeinander auf, von welchem Förderbereich wird zum anderen gewechselt? Gibt es hier spezifische Muster, sind diese vergleichbar zwischen den Kindern oder zeigen verschiedene Gruppen von Kindern in sich ähnliche aber zwischen den Gruppen unterschiedliche Muster? Gibt es hier Zusammenhänge mit Alter, nonverbaler und verbaler Entwicklung sowie autistischer Kernsymptomatik und anderen Verhaltensweisen?

  - Diskussionspunkt: bisher liegt Fokus auf Ketten von Zielen, diese Untersuchung würde Fokus auf Transitionen von Zielen erweitern

## modelling details

	- Ketten von zielen herausarbeiten
		- Mindestens zweimal bis dreimal bearbeitete Ziele berücksichtigen
	- Outcome Ziel: 
		- Zielebene gemeistert (2)
		- Ausmaß Hilfestellung (2)
		- Zielanzahl
		- Wie lange Zielkette?
	- Pro Ziel und Kind geschaut, wie Lange an Ziel gearbeitet und mastery
	- Anhand der Ketten und Qualität der ketten Clustern

## progress log 

### January 2026

	1. Welche Faktoren beeinflussen Beginn der Arbeit an Zielen? - see [moderating factors](#moderating-factors)
		- Einfluss Alter, Geschlecht, Parent education, Migrationsstatus, Corona, IQ, SES
		- -> age, IQ, and lockdown haben einen Einfluss
	2. Cluster der Therpapieziele zum Einstieg
		- 5 clusters that retain a language cluster
		- covariates and demographics have no influence on starting cluster
	3. Session Verlauf - anhand der Kettenmodellierung
		a. Ketten auf Ebene der Ziele (REMAINS THE CURRENT FOCUS)
		b. Ketten auf Ebene der Entwicklungsbereiche  (developmental area)
		c. Ketten auf Ebene der Einzelziele (Expertlevel)

### March 2026 - descriptive analysis

  1. Anzahl der Therapiesitzungen bis an Therapiezielen überhaupt gearbeitet werden kann
    - mean = 7.21 sessions, SD = 8.31 session, min-max = 2-60
  2. Zeit bis an Therapiezielen überhaupt gearbeitet werden kann
    - mean = 32 days, SD = 40.3 days, min-max = 1-253

### May 2026

  1. added sample description after final sample definition, see [descriptive table](#descriptive-table)
  2. excluded participants with less than 10 sessions (n=2)
  3. improved markdown visualization
  4. removed chains length < 2
  5. chain analysis restricted to three most common dev areas
  6. alternative models to LMM explored --> Bayesian model 
  

## TODO

  - model transitions of aims between chains
  - discuss: exclude low occurances aims (focus on prevalent aims)
  - discuss: exclude chains of length < 2
		
# raw data load

- loads the data

```{r load data}

#session data --> multiple rows per participant per session
project_folder<-"C:/Users/nico/Nextcloud/data_AFFIP/rct_data"
session_path<-paste0(project_folder,'/export_session_20230728.csv')
df_session<-read.csv(session_path)

#therapy data --> one row per participant per session
therapy_path<-paste0(project_folder,'/export_therapy_20230728.csv') #therapy session info - per session
df_therapy<-read.csv(therapy_path)

#gender and age in months
screen_path<-paste0(project_folder,'/export_screen_20230728.csv')
df_screen<-read.csv(screen_path) #master merge file
df_screen<-df_screen[,names(df_screen) %in% c('centno','scrnno','gender','agedemo','age_y','age_m')]

#IQ
wppsi_path<-paste0(project_folder,'/export_wppsi_20240916.csv')
df_wppsi<-read.csv(wppsi_path)
bayley_path<-paste0(project_folder,'/export_bayley_20240916.csv')
df_bayley<-read.csv(bayley_path)

#migration info
rct_path<-paste0(project_folder,'/export_rct_20230728.csv') 
df_rct<-read.csv(rct_path)

#socioeconomic status
ses_path<-paste0(project_folder,'/export_ses_20230728.csv') #socioeconomic status
df_ses<-read.csv(ses_path)


```

## IQ definition

- based on A-FFIP IQ definitions 2024 by Prof. Freitag 
- IQ is available as either Bayley or WPPSI

```{r define IQ}

#based on A-FFIP data preprocessing script


#select relevant data
df_wppsi <- df_wppsi[,names(df_wppsi) %in% c('centno', 'scrnno', 't', 'wppsi_ver','wppsidate',
                                             'vtiq', 'htiq', 'vgiq', 'asiq', 'gsiq', 'ta_wppsi_mean',
                                             'develop_age_w','verbal_iqdq_w','nonverbal_iqdq_w',
                                             'wppsi_ver','wppsidate_d','wppsidate_m','wppsidate_y')]

df_bayley<-df_bayley[,names(df_bayley) %in% c('centno','scrnno','t','baydate','develop_age',
                                              'verbal_iqdq','nonverbal_iqdq',
                                              'cognea','rezea','exprea')]

#merge data
df_iq<-merge(df_wppsi,df_bayley,by=c('centno','scrnno','t'),all.x=T,all.y=T)

#merge chronological age
df_age<-df_screen[,names(df_screen) %in% c('centno','scrnno','agedemo')]
names(df_age)[3]<-'chronage'
df_iq<-merge(df_iq,df_age,by=c('centno','scrnno'))

#Testalter (historisch)
#IQ estimate as test age /developmental age * 100
#df_iq$test_age<-with(df_iq,ifelse(is.na(ta_wppsi_mean),develop_age,ta_wppsi_mean))
df_iq$test_age<-with(df_iq,ifelse(is.na(develop_age_w),develop_age,develop_age_w))

#langugae development - not aggregated
df_iq$lang_dev<-df_iq$develop_age
df_iq$lang_dev_w<-df_iq$chronage / 100 * df_iq$asiq

#non-verbal IQ
nonverbal_iq<-df_iq$cognea / df_iq$chronage * 100
nonverbal_iq_w<-df_iq$htiq
df_iq$nonverbal_iq<-ifelse(is.na(nonverbal_iq),nonverbal_iq_w,nonverbal_iq)

#verbal IQ 
verbal_iq<-((df_iq$exprea + df_iq$rezea) / 2) / df_iq$chronage * 100
verbal_iq_w<-df_iq$vtiq
df_iq$verbal_iq<-ifelse(is.na(verbal_iq),verbal_iq_w,verbal_iq)

#full IQ 
full_iq<-((df_iq$exprea + df_iq$rezea + df_iq$cognea) / 3) / df_iq$chronage * 100
full_iq_w<-df_iq$gsiq
df_iq$full_iq<-ifelse(is.na(full_iq),full_iq_w,full_iq)

#full IQ - if nonverbal was measured with WPPSI_III and verbal with Bayley-III
# --> recovers 12 full scale IQs (!)
boolean_mixed<-!is.na(verbal_iq) & !is.na(nonverbal_iq_w)
df_iq$full_iq[boolean_mixed]<-
  ((df_iq$exprea[boolean_mixed]/df_iq$chronage[boolean_mixed] * 100) +
  (df_iq$rezea[boolean_mixed]/df_iq$chronage[boolean_mixed] * 100) +
  df_iq$htiq[boolean_mixed]) / 3
  
#developmental age
develop_age<-(df_iq$exprea + df_iq$rezea + df_iq$cognea) / 3
develop_age_w<-df_iq$chronage / 100 * df_iq$gsiq
df_iq$develop_age<-develop_age
df_iq$develop_age_w<-develop_age_w

```

# raw data overview

- different data frames for:
  - per session properties: df_therapy
  - per sim properties (multiple per session): df_session

## session properties

- session were with caregiver (thcg),
- coding is derived from: "A-FFIP Studie Case Report Form - Therapiedokumentation"
- analysis should be limited to: 
  - Bezugsperson anwesend = 0 (ohne Bezugsperson, variable: thcg)
  - Schwerpunkt = 2 (kindzentriert, variable: focus)
  - Setting = 1 (Einzeltherapie, variable: set)
  - Diagnostiktermin = 0 (kein Diagnostiktermin, variable: diag)

```{r variable description}

#Bezugsperson anwesen
table(df_therapy$thcg) %>%
  kbl(caption = "thcg: Bezugsperson anwesend (0 = no, 1=yes)") %>%
  kable_classic(full_width = T, html_font = "Cambria") 
# --> inclusion only with 0

#Schwerpunkt
table(df_therapy$focus) %>%
  kbl(caption = "focus: Therapieschwerpunkt (1 = Elternarbeit, 2 = Kind, 3 = Austausch / Umfeld)") %>%
  kable_classic(full_width = T, html_font = "Cambria") 
# 1 = Elternarbeit
# 2 = Kind
# 3 = Austausch / Umfeld

#Setting
table(df_therapy$set) %>%
  kbl(caption = "set: Setting (1 = Einzel, 2 = Gruppe, 3 = Gespräch, 4 = Hospitation Kindergarten") %>%
  kable_classic(full_width = T, html_font = "Cambria") 
# 1 = Einzel
# 2 = Gruppe
# 3 = Gespräch
# 4 = Hospitation Kindergarten

table(df_therapy$thcg,df_therapy$focus,df_therapy$set) %>%
  kbl(caption = "Bezugsperson x Schwerpunkt x Setting",
      col.names = c('Bezugsperson','Schwerpunkt','Setting','N')) %>%
  kable_classic(full_width = T, html_font = "Cambria") 

#Diagnostiktermin 
table(df_therapy$diag) %>%
  kbl(caption = "diag: Diagnostiktermin (0 = nein)") %>%
  kable_classic(full_width = T, html_font = "Cambria") 
# 0 = kein Diagnostiktermin

```

## aim properties

- multiple aims per session
- is a different data frame in the raw data and needs to me merged with session properties
- number of aims within the session
- there is no priority or rank of aims

```{r aims_per_session}

#Ziel Nummer
table(df_session$aim)  %>%
  kbl(caption = "aim: Mindestanzhal von Zielen pro Sitzung (kummulativ)") %>%
  kable_classic(full_width = T, html_font = "Cambria") 
##-->  aim number within the session (3 = third of three aims)
## 	
```

## aim practice

- only consider aims that were actually trained (practex), additional inclusion criterion

```{r aim_practice}

#Übung durchgeführt
table(df_session$practex) %>%
  kbl(caption = "practex: Übung durchgeführt") %>%
  kable_classic(full_width = T, html_font = "Cambria")

#if Übung durchgeführt, then: Übung variiert
table(df_session$practvar)
## --> currentlky not relevant

```

## outcome variables

- help (0-2) and achieve (0-2) define a core outcome of the analysis
- behave (0 versus 1) could be a secondary outcome variable

```{r add_vars}

#Ausmaß Hilfestellung: 0 = ohne Hilfe nicht geschafft, 1 = teilweise selbstständig, 2 = selbstständig
table(df_session$help) %>%
  kbl(caption = "help: Ausmaß Hilfestellung (0 = Hilfe notwendig, 1 = teilweise selbstständig, 2 = selbstständig)") %>%
  kable_classic(full_width = T, html_font = "Cambria") 

#Zielebene gemeistert - 0= nein, 1 = teilweise, 2 = vollständig
table(df_session$achieve) %>%
  kbl(caption = "achieve: Zielebene gemeistert (0 = nein, 1 = teilweise, 2 = vollständig)") %>%
  kable_classic(full_width = T, html_font = "Cambria") 

#herausfordernde Verhaltensweisen gezeigt - 0 = nein, 1 = ja
table(df_session$behave) %>%
  kbl(caption = "behave: Herausfordernde Verhaltensweisen (0 = nein, 1 = ja)") %>%
  kable_classic(full_width = T, html_font = "Cambria") 

```

# data set definition

- combines session and therapy data, adding demographics, and select sessions based on criteria

## merge data

- merges df_session (session data, p.12ff) and df_therapy (therapy data, p1-12)
- df_session: row per aim per session (n=27579)
- df_therapy: row per session (n=4731)
- adds an ID variable based on centno + scrnno
- result: df_trial

```{r merge data}

df_trial<-merge(df_session, df_therapy, 
            by.x = c("centno", "scrnno", "sessno"), 
            by.y= c("centno", "scrnno", "thsessno"))

#create IDs variable
df_trial$id<-paste(df_trial$centno,df_trial$scrnno,sep='.')

```

## add demographics

- IQ 
- gender 
- age 
- migration info 
- parental education

```{r adding demographic data}

#adds agedemo, gender
df_trial<-merge(df_trial,df_screen, by = c("centno", "scrnno"))

#adds iq
df_iq<-df_iq[df_iq$t==1,names(df_iq)!='t'] #at earliest timepoint
df_trial<-merge(df_trial,df_iq, by = c("centno", "scrnno"))

#migration info
df_rct<-df_rct[,!(names(df_rct) %in% c('bdate_m','bdate_y'))]
df_rct<-df_rct[,names(df_rct)!='t']
df_trial<-merge(df_trial,df_rct, by = c("centno", "scrnno"), all.x = T)

#parental education and SES
selected_vars<-c("centno","scrnno","ses_edumean_t1","maxschool_t1")
df_ses<-df_ses[,names(df_ses) %in% selected_vars]
df_trial<-merge(df_trial,df_ses, by = c("centno", "scrnno"))

```

## included sessions

- retain only data for session that:
  - without caregiver (df_trial$thcg==0)
  - focused on child (df_trial$focus==2)
  - Einzeltherapie (df_trial$set==1)
  - no diagnostic assessment (df_trial$diag==0)
- retain only data for aims that:
  - actually trained aims (df_trail$practex==1)

```{r define_sample}

# total_aims<-nrow(df_trial)
# print(paste0('total aims in raw data: ',total_aims))
# 
# print('aims without caregiver (thcg==0):')
# table(df_trial$thcg==0)
# 
# print('aims focused on child (focus==2):')
# table(df_trial$focus==2)
# 
# print('aims in single child setting (set==1):')
# table(df_trial$set==1)
# 
# print('aims outside diagnosic sessions (diag==0):')
# table(df_trial$diag==0)
# 
# print('aims actually practiced (practex==1):')
# table(df_trial$practex==1)
# 
# #selection step
# df_trial<-df_trial[df_trial$thcg==0 & df_trial$focus==2 & df_trial$set==1 & df_trial$diag==0 & df_trial$practex==1,]
# 
# print(paste0('retained data percentage: ',round(nrow(df_trial)/total_aims,2)))
# print(paste0('retained aims: ',round(nrow(df_trial))))

total_aims <- nrow(df_trial)

# Step 1: diag == 0
keep1 <- !is.na(df_trial$diag) & df_trial$diag == 0
step1_excluded <- sum(!keep1)
df_step1 <- df_trial[keep1, ]
step1_remaining <- nrow(df_step1)

# Step 2: focus == 2
keep2 <- !is.na(df_step1$focus) & df_step1$focus == 2
step2_excluded <- sum(!keep2)
df_step2 <- df_step1[keep2, ]
step2_remaining <- nrow(df_step2)

# Step 3: thcg == 0
keep3 <- !is.na(df_step2$thcg) & df_step2$thcg == 0
step3_excluded <- sum(!keep3)
df_step3 <- df_step2[keep3, ]
step3_remaining <- nrow(df_step3)

# Step 4: set == 1
keep4 <- !is.na(df_step3$set) & df_step3$set == 1
step4_excluded <- sum(!keep4)
df_step4 <- df_step3[keep4, ]
step4_remaining <- nrow(df_step4)

# Step 5: practex == 1
keep5 <- !is.na(df_step4$practex) & df_step4$practex == 1
step5_excluded <- sum(!keep5)
df_step5 <- df_step4[keep5, ]
step5_remaining <- nrow(df_step5)

# Final selected dataset
df_trial_selected <- df_step5

print(paste0("Total aims in raw data: ", total_aims))
print(paste0("Retained data percentage: ", round(nrow(df_trial_selected) / total_aims, 2)))
print(paste0("Retained aims: ", nrow(df_trial_selected)))

#flowchart
grViz(paste0("
digraph flowchart {
  graph [layout = dot, rankdir = TB]
  node [shape = box, style = rounded, fontname = Helvetica]
  
  main0 [label = 'Raw data\nN = ", total_aims, "', style=filled, fillcolor=LightBlue]
  main1 [label = 'diag == 0\nTherapy sessions\nN = ", step1_remaining, "', style=filled, fillcolor=LightBlue]
  main2 [label = 'focus == 2\nChild-focused\nN = ", step2_remaining, "', style=filled, fillcolor=LightBlue]
  main3 [label = 'thcg == 0\nNo caregiver present\nN = ", step3_remaining, "', style=filled, fillcolor=LightBlue]
  main4 [label = 'set == 1\nSingle-child setting\nN = ", step4_remaining, "', style=filled, fillcolor=LightBlue]
  main5 [label = 'practex == 1\nActually practiced aims\nFinal N = ", step5_remaining, "', style=filled, fillcolor=Green]

  ex1 [label = 'Excluded\ndiag != 0 or NA\nn = ", step1_excluded, "', style=filled, fillcolor=MistyRose]
  ex2 [label = 'Excluded\nfocus != 2 or NA\nn = ", step2_excluded, "', style=filled, fillcolor=MistyRose]
  ex3 [label = 'Excluded\nthcg != 0 or NA\nn = ", step3_excluded, "', style=filled, fillcolor=MistyRose]
  ex4 [label = 'Excluded\nset != 1 or NA\nn = ", step4_excluded, "', style=filled, fillcolor=MistyRose]
  ex5 [label = 'Excluded\npractex != 1 or NA\nn = ", step5_excluded, "', style=filled, fillcolor=MistyRose]

  main0 -> main1
  main1 -> main2
  main2 -> main3
  main3 -> main4
  main4 -> main5

  main0 -> ex1 [dir=none]
  main1 -> ex2 [dir=none]
  main2 -> ex3 [dir=none]
  main3 -> ex4 [dir=none]
  main4 -> ex5 [dir=none]
}
"))

#retain data
df_trial<-df_trial_selected

```

## pandemic break

- first shutdown in Germany: 22/03/2020 - 04/05/2020
- calculate a corona state based on lockdown onset (after_lockdown=T == session after 22/03/20)

```{r pandemic_break}

  #lockdown as officially defined:
  lockdown_start_germany<-as.Date("22/03/20", format = "%d/%m/%y")
  lockdown_end_germany<-as.Date("04/05/20", format = "%d/%m/%y")

  #across all
  df_trial$sdate_<-as.Date(df_trial$sdate_, format = "%d/%m/%y")
  ggplot(df_trial,aes(sdate_,sessno,color=id))+geom_point()+
    geom_vline(xintercept=lockdown_start_germany,color='red')+
    geom_vline(xintercept=lockdown_end_germany,color='red')+
    theme(legend.position = "none")+
    labs(x='year',y='session number',title='sessions over duration of the study',subtitle='each color represents a participant')
    
hist(df_trial$sdate_,150,main='sessions across time of study')

#define variable
df_trial$after_lockdown<-ifelse(df_trial$sdate_>lockdown_start_germany,T,F)
  
print(paste0('proportion of therapy session after first lockdown (22/03/20): ',
             round(nrow(df_trial[df_trial$after_lockdown==T,])/nrow(df_trial),2)))
  
```

# sample description

## exclude participants

- exclude participants with less than 10 session 
  - only one session (id = 1.32)
  - only 6 sessions (id = 4.23)

```{r}

df_trial<-df_trial[!df_trial$id=='1.32',]
df_trial<-df_trial[!df_trial$id=='4.23',]


```

## participants & sessions overview

- shows participants ids (centno.scrnno)
- sessions by participant
- define an ID variable for df_trial

```{r number_of_participants}

#participants
print('number of participants:')
length(table(interaction(df_trial$centno,df_trial$scrnno,drop=T)))

#session by participant
sessions_by_participant<-as.numeric(
  by(df_trial$sessno,interaction(df_trial$centno,df_trial$scrnno,drop=T),function(x){length(unique(x))}))
hist(sessions_by_participant,main = 'sessions by participant')

participant_IDs<-names(table(interaction(df_trial$centno,df_trial$scrnno,drop=T)))

df_ids<-data.frame(participant_IDs,sessions_by_participant)
df_ids<-df_ids[order(df_ids$sessions_by_participant),]

df_ids %>%
  kbl(caption = "Sessions by participant") %>%
  kable_classic(full_width = T, html_font = "Cambria") 

```

## sample description table {#descriptive-table}

```{r sample_description_table}

description_variables<-c('centno','scrnno','agedemo','gender',
                         'nonverbal_iq','verbal_iq','full_iq',
                         'migration','language','ses_edumean_t1','maxschool_t1')

#create data frame with unique ids
df_sample<-df_trial[!duplicated(interaction(df_trial$centno,df_trial$scrnno,drop=T)),names(df_trial) %in% description_variables]

#table for categorical variables
gender<-table(df_sample$gender)
migration<-table(df_sample$migration)
language<-table(df_sample$language)

df_sample_categorical<-data.frame(rbind(gender,migration,language))
row_names<-c('gender (0 = male, 1=female)','migration history of parents (0=yes, 1=no)',
             'language spoken at home (0=German, 1=other)')
df_sample_categorical<-cbind(row_names,df_sample_categorical)
names(df_sample_categorical)<-c('variable','0','1')

df_sample_categorical %>%
  kbl(caption = "Table 1A. sample description for categorical variables",row.names = F) %>%
  kable_classic(full_width = T, html_font = "Cambria") 


#table for numeric variables
df_sample_numeric<-rbind(
  psych::describe(df_ids$sessions_by_participant), #note: utilizes different data set
  psych::describe(df_sample$agedemo),
  psych::describe(df_sample$nonverbal_iq),
  psych::describe(df_sample$verbal_iq),
  psych::describe(df_sample$full_iq),
  psych::describe(df_sample$ses_edumean_t1),
  psych::describe(df_sample$maxschool_t1)
)

df_sample_numeric<-df_sample_numeric[,c('n','mean','sd','min','max')]
df_sample_numeric<-round(df_sample_numeric,2)

row.names(df_sample_numeric)<-c('number of sessions','age (in m)','nonverbal IQ','verbal IQ','full IQ',
                                'parents education (mean)','parent education (max)')

df_sample_numeric %>%
  kbl(caption = "Table 1B. Sample description for numeric variables") %>%
  kable_classic(full_width = F, html_font = "Cambria") %>%
  footnote(general = c("parent education: 0 = no school (ISCED 0)","1 = school not completed (ISCED 1)","2 = completed special school (ISCED 1)","3 = Hauptschulabschluss (ISCED 2A)",'4 = Realschulabschluss/Berufsfachschule (ISCED3B)',"5 = Hochschulreife/Abitur (ISCED 3A)"))

```

# therapy aims overview

- can be more than one per session
- encoded by "aimno" variable (e.g. 10.6.2.1)
  - first or two numbers define devleopmental area (e.g. 9, 10.2-10.6)
  - first three numbers define specific aim (e.g. 10.6.2) --> is our target
  - last number defines expert level (1,2,3) --> rather as outcome variable
  
## developmental area

- Entwicklungsbereiche (9.2 - 10.6)

```{r developmental area}

## Ziele overall
#table(df_trial$aimno)

#create data
dev_area<-substr(df_trial$aimno,1,8)
dev_area<-ifelse(nchar(dev_area)==5,substr(dev_area,1,1),dev_area)
dev_area<-ifelse(nchar(dev_area)==8,substr(dev_area,1,4),dev_area)
dev_area<-ifelse(grepl("A",dev_area),'A',dev_area)
dev_area<-ifelse(grepl("B",dev_area),'B',dev_area)
dev_area<-ifelse(grepl("-8",dev_area),NA,dev_area)

dev_area_char<-dev_area
dev_area<-as.factor(dev_area)
levels(dev_area)<-c('communication & speech (10.2)','interaction & play (10.3)','emotion (10.4)','cognitive ability (10.5)','daily living skills (10.6)','basic abilities (9)','introduction (A)','reduction of unwanted behaviors (B)')

df_trial$dev_area<-dev_area
df_trial$dev_area_char<-dev_area_char

#visualize data
dev_area_table<-data.frame(sort(table(df_trial$dev_area),decreasing = TRUE))
dev_area_table_char<-data.frame(sort(table(df_trial$dev_area_char),decreasing = TRUE))
dev_area_table$dev_area_char<-dev_area_table_char$Var1
dev_area_table<-dev_area_table[,c(1,3,2)]

dev_area_table %>%
  kbl(caption = "developmental areas across sessions") %>%
  kable_classic(full_width = T, html_font = "Cambria") 

ggplot(df_trial,aes(x=dev_area,fill=dev_area))+geom_bar()+theme(axis.text.x=element_blank())

```

## aims across sessions

- an welchen zielen wurde gearbeitet?

```{r therapy aims}

## Ziele overall
#table(df_trial$aimno)

aims<-substr(df_trial$aimno,1,8)
aims<-ifelse(nchar(aims)==8,substr(aims,1,6),aims)
aims<-ifelse(nchar(aims)==5,substr(aims,1,3),aims)
aims<-ifelse(grepl("-8",aims),NA,aims)

aims_rank<-aims
aims_rank<-ifelse(grepl('10',aims_rank),substr(aims_rank,6,6),aims_rank)
aims_rank<-ifelse(grepl('9',aims_rank),substr(aims_rank,3,3),aims_rank)
aims_rank<-ifelse(grepl('A',aims_rank),NA,aims_rank)
aims_rank<-ifelse(grepl('B',aims_rank),NA,aims_rank)
df_trial$aims_rank<-aims_rank

aims_char<-aims
aims<-as.factor(aims)
levels(aims)<-c('eye contact','passive vocabulary','gesture','expressive language (supported)',
                'expressive language extended','questions & answers','making contact',
                
                'shared attention','exchange','sharing','non-interactive play','simple shared play',
                'rule-based play','symbolic play','role play',
                
                'emotion recognition','emotion expression','emotion regulation',
                
                'classification','characteristics/opposites','memory',
                'drawing','quantitative understanding','theory of mind',
                'common knowledge',
                
                'daily routines','toilet training',
                'problem solving','awareness of danger','spatial awareness','temporal orientation',
                
                'attention control','gaze following',
                'imitation','representation','action planing','self awareness',
                
                'testing reinforcement','building therapeutic alliance','situational control','separation from caregiver',
                
                'aggression','stereotypic behavior','alternatives','anxiety','avoidance'
                
                )

df_trial$aims<-aims
df_trial$aims_char<-aims_char

aims_table<-data.frame(sort(table(df_trial$aims),decreasing = TRUE))
aims_table_char<-data.frame(sort(table(df_trial$aims_char),decreasing = TRUE))
aims_table$aims_char<-aims_table_char$Var1
aims_table<-aims_table[,c(1,3,2)]

  aims_table %>%
  kbl(caption = "occurance of aims across sessions") %>%
  kable_classic(full_width = F, html_font = "Cambria") 

#commented out as aim rank is not relevant  
  
# aims_table_byrank<-as.matrix(table(df_trial$dev_area,df_trial$aims_rank))
# 
#   aims_table_byrank %>%
#   kbl(caption = "aim ranks by developmental area across sessions") %>%
#   kable_classic(full_width = F, html_font = "Cambria") 

```

# first aims 

- extract first session per participant´and according aims
- remove introduction sessions
- the number of aims in the first session can be more than one per child 

```{r create data of first session aims}

#create data
df_firstaims<-df_trial[df_trial$dev_area!='introduction (A)',]
df_firstaims<-split(df_firstaims,df_firstaims$id)
df_firstaims<-lapply(df_firstaims,function(x){
  earliest_session<-min(x$sessno)
  x<-x[x$sessno==earliest_session,]
  return(x)
})
df_firstaims<-rbindlist(df_firstaims)

```

## first aims overview

- top 3: eye contact (10.2.1), imitation (9.4), gaze following (9.3)

```{r first aims across number of aims}

#table overview
aims_table<-sort(table(droplevels(df_firstaims$aims)),decreasing = TRUE)
aims_table<-data.frame(aims_table)

aims_char_table<-sort(table(df_firstaims$aims_char),decreasing = TRUE)
aims_table<-data.frame(names(aims_char_table),aims_table)
names(aims_table)<-c('code','aim','occurances')

  aims_table %>%
  kbl(caption = "aims in the first session after introdcution") %>%
  kable_classic(full_width = F, html_font = "Cambria") 

#aims by participant
table_aimspart<-by(df_firstaims$aims,df_firstaims$id,function(x){table(droplevels(x))})

table_aimspart<-sapply(table_aimspart,dimnames)
max_length<-max(sapply(table_aimspart,length))
table_aimspart<-lapply(table_aimspart,function(x){
  current_length<-length(x)
  new_vector<-rep(NA,max_length)
  new_vector[1:current_length]<-x
  return(new_vector)})

table_aimspart<-rbind.data.frame(table_aimspart)

table_aimspart<-data.frame(t(table_aimspart))
names(table_aimspart)<-paste('goal',1:length(names(table_aimspart)))
row.names(table_aimspart)<-substr(row.names(table_aimspart),2,nchar(row.names(table_aimspart)))

print('discussion point: Factor in the number of aims per participant?')

table_aimspart[order(table_aimspart[,1]),] %>%
  kbl(caption = "aims in the first session after introdcution by participant") %>%
  kable_classic(full_width = F, html_font = "Cambria") 

```

## time till first aims

```{r, time span to first session}

#extract date of first session
df_firstsession<-df_therapy[df_therapy$thsessno==1,]
df_firstsession<-df_firstsession[,names(df_firstsession) %in% c('centno','scrnno','thdate_')]
names(df_firstsession)<-c('centno','scrnno','date_first_session')

#add - adds df_firstaims that already removed A/introduction codings
df_firstaims<-merge(df_firstaims,df_firstsession,by =c("centno", "scrnno"))

#convert to date
df_firstaims$sdate_<- as.Date(df_firstaims$sdate_, format = "%d/%m/%y")
df_firstaims$date_first_session<- as.Date(df_firstaims$date_first_session, format = "%d/%m/%y")

#select only one row per participant
df_firstaims_uniqueID<-df_firstaims[!duplicated(df_firstaims$id),]

time_to_first_session<-data.frame(df_firstaims_uniqueID$sdate_-df_firstaims_uniqueID$date_first_session)
names(time_to_first_session)<-'time_in_d'

# table(time_to_first_session) %>%
#    kbl(caption = "time in days from introduction to session with first aims") %>%
#   kable_classic(full_width = T, html_font = "Cambria") 
# 
# ggplot(time_to_first_session,aes(x=time_in_d))+geom_bar()+xlab('time in days to first session with aims')


df_date_first_session<-data.frame(df_firstaims_uniqueID$id,time_to_first_session$time_in_d,
                                  df_firstaims_uniqueID$sessno,df_firstaims_uniqueID$sdate_)

df_date_first_session$days_by_session<-round(df_date_first_session$time_to_first_session.time_in_d/
                                               df_date_first_session$df_firstaims_uniqueID.sessno,2)

names(df_date_first_session)<-c('id','days to first session','session number','date of first session','days / session')
df_date_first_session<-df_date_first_session[order(df_date_first_session$'days to first session'),]

#descriptive data
summary(df_date_first_session$`days to first session`)
sd(df_date_first_session$`days to first session`)

#table
df_date_first_session %>%
   kbl(caption = "first session with aims: date and time by IDs", row.names = FALSE) %>%
   kable_classic(full_width = T, html_font = "Cambria")


```

## sessions till first aims

- Berechnung der Anzahl der Therapiesitzungen, die Kind gebraucht hat, bis an Therapiezielen überhaupt gearbeitet werden kann - Kriterium: Anzahl der „A“-Sitzungen zu Beginn der Therapie“

```{r first session as therapy session}

psych::describe(df_firstaims$sessno[!duplicated(df_firstaims$id)])

#session number after introduction   
hist(df_firstaims$sessno[!duplicated(df_firstaims$id)],
     main = 'session number when first aims',20)  

table(df_firstaims$sessno[!duplicated(df_firstaims$id)]) %>%
  kbl(caption = "Session number after introduction (after 'A')") %>%
  kable_classic(full_width = T, html_font = "Cambria") 

```

### outlier investigation

- id 2.4 has 60 sessions till first aims
- caregiver was present during first 60 sessions
- has 30 sessions without caregiver after that - leave in analysis

```{r outlier_analysis first session}

outlier_sessions<-df_session[df_session$centno=='2' & df_session$scrnno=='4',]
outlier_sessions$sdate_<-as.Date(outlier_sessions$sdate_, format = "%d/%m/%y")
ggplot(outlier_sessions,aes(sdate_,sessno,color=aimno))+geom_point()+labs(title= 'outlier participant concerning duration of introduction',subtitle = 'color encodes different aims that were worked on for this participant while caregiver present')

outlier_sessions<-outlier_sessions[order(outlier_sessions$sessno),]

outlier_sessions2<-df_therapy[df_therapy$centno=='2' & df_therapy$scrnno=='4',]
table(outlier_sessions2$thcg)

```

## moderating factors {#moderating-factors}

- which factors influence the start of working on first aims
- consider age, gender, parental education, migration status, corona, IQ, SES

- for non-parametric testing: age, IQ, and lockdown have an effect on session number
  - age: spearman correlation: rho = -0.31, p = .014
  - IQ: spearman correlation: rho = -0.26, p = .045
  - lockdown: session before lockdown: m/SD = 7.03/10.45, session after lockdown: m/SD = 7.57/5.18, p = .47

--> consider those as covariates in testing?

```{r moderating factors on first aim}

#select only one row per participant
df_firstaims_uniqueID<-df_firstaims[!duplicated(df_firstaims$id),]

#nonparametric testing - categorical variables
print('gender:')
with(df_firstaims_uniqueID,kruskal.test(sessno ~ as.factor(gender)))
print('migration status:')
with(df_firstaims_uniqueID,kruskal.test(sessno ~ as.factor(migration), data = df_firstaims_uniqueID))
print('parental education:')
with(df_firstaims_uniqueID,kruskal.test(sessno ~ as.factor(maxschool_t1), data = df_firstaims_uniqueID))
print('lockdown status:')
with(df_firstaims_uniqueID,kruskal.test(sessno ~ as.factor(after_lockdown), data = df_firstaims_uniqueID))

#--> significant influence of lockdown - sessions after lockdown take longer to reach aims
with(df_firstaims_uniqueID,by(sessno,after_lockdown,psych::describe))

#nonparametric testing - continous variables
print('age:')
with(df_firstaims_uniqueID, cor.test(agedemo, sessno, method = "spearman",exact=F))
print('IQ:')
with(df_firstaims_uniqueID,cor.test(full_iq, sessno, method = "spearman",exact=F))
print('SES:')
with(df_firstaims_uniqueID,cor.test(ses_edumean_t1, sessno, method = "spearman",exact=F))

```

## clustering first aims {#clustering-aims}

- Cluster method: Gower Distance Hierarchical Clustering
- consider the first 7 goals of every participant at first session
- considers missing data (not all participants have equal number of aims)
- categorical data type

```{r first aim clustering, fig.width=10, fig.height=6}

# Cut into clusters (e.g., 3 clusters)
number_of_clusters<-5
print(paste('assumed number of clusters:',number_of_clusters))

#translate 
table_aimspart_factor<-lapply(table_aimspart,as.factor)
table_aimspart_factor<-data.frame(table_aimspart_factor[1],
                                  table_aimspart_factor[2],
                                  table_aimspart_factor[3],
                                  table_aimspart_factor[4],
                                  table_aimspart_factor[5],
                                  table_aimspart_factor[6],
                                  table_aimspart_factor[7])

rownames(table_aimspart_factor)<-row.names(table_aimspart)


# Compute Gower distance
gower_dist <- daisy(table_aimspart_factor, metric = "gower")
print('clustering metric - gower distance between pairwise comparisons:')
table(round(gower_dist,2))

# Perform hierarchical clustering
hc <- hclust(gower_dist, method = "ward.D2")
clusters <- cutree(hc, k = number_of_clusters)

# Add cluster labels to your data
table_aimspart_factor$start_cluster <- clusters

print('5 clusters combine language aims better into distinct cluster')

# table_aimspart_factor[order(table_aimspart_factor[,8]),] %>%
#   kbl(caption = "aims in the first session and starting CLUSTERS after introduction by participant") %>%
#   kable_classic(full_width = F, html_font = "Cambria") 

# Labels for dendrogram
dendro_labels <- paste0(rownames(table_aimspart_factor), " [Cluster ", clusters, "]")

# visualize cluster
plot(hc, labels = dendro_labels, main = "Dendrogram of Hierarchical Clustering")
rect.hclust(hc, k = number_of_clusters, border = "red")  # Draw boxes for clusters

table(table_aimspart_factor$start_cluster)

#add clusters to data
id<-row.names(table_aimspart_factor)
start_cluster<-table_aimspart_factor$start_cluster
df_cluster<-data.frame(id,start_cluster)
df_firstaims<-merge(df_firstaims,df_cluster,by='id')
df_firstaims$start_cluster<-as.factor(df_firstaims$start_cluster)

df_firstaims_uniqueID<-data.frame(df_firstaims[!duplicated(df_firstaims$id),])
df_firstaims_uniqueID<-df_firstaims_uniqueID[,names(df_firstaims_uniqueID) %in% c('id','start_cluster')]
df_trial<-merge(df_trial,df_firstaims_uniqueID,by='id')

```

### describe clusters

- 5 most common aims in different clusters for first therapy session aims
- labels the clusters 

```{r describe initial clusters}

#show 5 most common aims in different clusters
cluster_aims<-by(table_aimspart_factor[,1:7],table_aimspart_factor$start_cluster,
   function(x){sort(table(droplevels(unlist(x))),decreasing=T)[1:5]})
cluster_aims<-lapply(cluster_aims,data.frame)
cluster_aims<-do.call(cbind,cluster_aims)
names(cluster_aims)<-rep(c('label','frequency'),number_of_clusters)

participants_in_clusters<-as.numeric(table(table_aimspart_factor$start_cluster))
header_label<-rep(2,number_of_clusters)
names(header_label)<-paste('cluster',1:number_of_clusters,'(n=',participants_in_clusters,')')

cluster_aims %>%
  kbl(caption = "5 most common aims in the first session by clusters") %>%
  kable_classic(full_width = F, html_font = "Cambria") %>%
  add_header_above(header_label)


```

### moderating clusters

- non-parametric Fisher's exact Test (categorical outcome = cluster, categorical predictor)
- no moderator had an influence

```{r moderating clusters}

#select only one row per participant
df_firstaims_uniqueID<-df_firstaims[!duplicated(df_firstaims$id),]
table(df_firstaims_uniqueID$start_cluster)

#nonparametric testing - categorical variables
print('gender:')
with(df_firstaims_uniqueID,table(start_cluster, gender))
with(df_firstaims_uniqueID,fisher.test(table(start_cluster, gender)))
print('migration status:')
with(df_firstaims_uniqueID,table(start_cluster, migration))
with(df_firstaims_uniqueID,fisher.test(table(start_cluster, migration)))
print('parental education:')
with(df_firstaims_uniqueID,table(start_cluster, maxschool_t1))
with(df_firstaims_uniqueID,fisher.test(table(start_cluster, maxschool_t1)))
print('lockdown status:')
with(df_firstaims_uniqueID,table(start_cluster, after_lockdown))
with(df_firstaims_uniqueID,fisher.test(table(start_cluster, after_lockdown)))

#--> significant influence of lockdown - sessions after lockdown take longer to reach aims
with(df_firstaims_uniqueID,by(sessno,after_lockdown,psych::describe))

#nonparametric testing - continous variables
print('age:')
with(df_firstaims_uniqueID, kruskal.test(agedemo ~ start_cluster))
print('IQ:')
with(df_firstaims_uniqueID, kruskal.test(full_iq ~ start_cluster))
print('SES:')
with(df_firstaims_uniqueID, kruskal.test(ses_edumean_t1 ~ start_cluster))

```


## therapy aims - expert level

- expert level defines a type of mastery of a specific aim
- last number defines expert level (1,2,3)
- can be utilized as outcome measure of training 

```{r}

expert_level<-substr(df_trial$aimno,5,8)
expert_level<-ifelse(nchar(expert_level)==4,substr(expert_level,4,4),expert_level)
expert_level<-ifelse(nchar(expert_level)==0,NA,expert_level)

expert_level<-as.factor(expert_level)
levels(expert_level)<-c('beginner','intermediate','expert')

df_trial$expert_level<-expert_level

with(df_trial,table(dev_area,expert_level))

ggplot(df_trial,aes(x=sessno,fill=expert_level))+geom_bar()+facet_wrap(~dev_area)


```

# chains of aims

- identifiy chains of consecutive working on specific aims 
- at least 2 consecutive sessions with one aim should be later considered

## chain functions

- code below define functions to extract chains of aims
- use "show" to examine code

```{r chain functions}

#function to extract outcome data for chains of a single aim within one participant
#outocme data includes: achieve, help
fun_chain_metrics<-function(single_chain){
#chain static info
single_chain_outcome<-single_chain[1,names(single_chain) %in% c('id','centno','scrnno','aims','aims_char',
                                                                'dev_area','dev_area_char','start_cluster',
                                                                'gender','agedemo','develop_age','verbal_iq',
                                                                'nonverbal_iq','full_iq','test_age',
                                                                'migration','maxschool_t1','ses_edumean_t1',
                                                                'after_lockdown')]


#chain characterization
chain_length<-nrow(single_chain)
chain_first_session<-min(single_chain$sessno)
chain_last_session<-max(single_chain$sessno)
chain_first_date<-min(single_chain$sdate_)
chain_last_date<-max(single_chain$sdate_)
chain_after_lockdown_percentage<-mean(single_chain$after_lockdown)
chain_behave_mean<-mean(single_chain$behave)
#chain outcome metrics
chain_achieve_mean<-mean(single_chain$achieve)
chain_achieve_first<-single_chain$achieve[single_chain$sessno==chain_first_session]
chain_achieve_last<-single_chain$achieve[single_chain$sessno==chain_last_session]
chain_achieve_0<-sum(single_chain$achieve==0)
chain_achieve_1<-sum(single_chain$achieve==1)
chain_achieve_2<-sum(single_chain$achieve==2)
chain_help_mean<-mean(single_chain$help)
chain_help_first<-single_chain$help[single_chain$sessno==chain_first_session]
chain_help_last<-single_chain$help[single_chain$sessno==chain_last_session]
chain_help_0<-sum(single_chain$help==0)
chain_help_1<-sum(single_chain$help==1)
chain_help_2<-sum(single_chain$help==2)
chain_expertlevel_beginner<-sum(single_chain$expert_level=='beginner')
chain_expertlevel_intermediate<-sum(single_chain$expert_level=='intermediate')
chain_expertlevel_expert<-sum(single_chain$expert_level=='expert')

single_chain_variables<-data.frame(
  #chain characterization
  chain_length,
  chain_first_session,
  chain_last_session,
  chain_first_date,
  chain_last_date,
  chain_after_lockdown_percentage,
  chain_behave_mean,
  #chain outcome metrics
  chain_achieve_mean,
  chain_achieve_first,
  chain_achieve_last,
  chain_achieve_0,
  chain_achieve_1,
  chain_achieve_2,
  chain_help_mean,
  chain_help_first,
  chain_help_last,
  chain_help_0,
  chain_help_1,
  chain_help_2,
  chain_expertlevel_beginner,
  chain_expertlevel_intermediate,
  chain_expertlevel_expert
  )

single_chain_outcome<-data.frame(single_chain_outcome,single_chain_variables)
return(single_chain_outcome)

}

#function that loops fun_chain_metrics across aims by participant
fun_chain_loop_ids_aims<-function(single_participant){
  
#testing:  
#single_participant<-list_trial[[22]]
  
#identify aims
all_aims<-names(table(droplevels(single_participant$aims)))

#identifies chains for participant - loops across all aims of a participant
single_participant_chains<-list()
for(i in 1:length(all_aims)){
single_participant_single_aim<-single_participant[single_participant$aims==all_aims[i],] 
# Identify breaks
group_id <- cumsum(c(TRUE, diff(single_participant_single_aim$sessno) != 1))
# Split into list of chains
chains <- split(single_participant_single_aim, group_id)
#apply fun_chain_metrics to each list element
chains_of_single_aim<-lapply(chains,fun_chain_metrics)
#return to data.frame
chains_of_single_aim<-rbindlist(chains_of_single_aim)
#retrieve number of chains for single aim
chains_of_single_aim$aim_chain_nr<-1:nrow(chains_of_single_aim)
single_participant_chains[[i]]<-chains_of_single_aim
}

names(single_participant_chains)<-all_aims #names list of chains by aims
single_participant_chains<-rbindlist(single_participant_chains) #turn list of chains to data.frame  
return(single_participant_chains)

}

```

## chain data extraction

- splits data by participants
- splits single participant by aims
- create chains for every participant (for loop)
- create mastery variable based on expert level
- remove chains with length < 2 (removes about 50% of chains; i.e. half of chains are 1 session chains)

```{r prepare data of chains}

#split by participant
list_trial<-split(df_trial,df_trial$id)
list_trial<-lapply(list_trial,function(x){x[order(x$sessno),]})

# applies fucntions above in loop (participants) of loops (aims per participant)
list_chains<-pblapply(list_trial,fun_chain_loop_ids_aims)
df_chains<-rbindlist(list_chains)

#create additional variable 
df_chains$chain_mastery_mean<-with(df_chains,
             ((chain_expertlevel_beginner*0)+(chain_expertlevel_intermediate*1)+(chain_expertlevel_expert*2))/
               (chain_expertlevel_beginner+chain_expertlevel_intermediate+chain_expertlevel_expert))

#remove chain length < 2 (removes 50% of data)
df_chains<-df_chains[df_chains$chain_length>1,]

```

## chain description

```{r chain description}

table(df_chains$chain_length) %>%
  kbl(caption = "total chain length") %>%
  kable_classic(full_width = F, html_font = "Cambria")

hist(with(df_chains,by(chain_length,aims,mean)),main = 'mean chain length')

table_chainlength<-data.frame(with(df_chains,by(chain_length,aims,function(x){round(mean(x),2)})))
chain_occurances<-as.numeric(with(df_chains,by(chain_length,aims,length)))
table_chainlength<-data.frame(table_chainlength,chain_occurances)

table_chainlength<-table_chainlength[order(table_chainlength[,2],decreasing = T),]
names(table_chainlength)<-c('mean chain length','chain occurances')

table_chainlength %>%
  kbl(caption = "mean chain length by aim") %>%
  kable_classic(full_width = F, html_font = "Cambria")

```

## chain visualization

```{r chain visualization}

ggplot(df_chains[!df_chains$dev_area_char %in% c('A','B')],
       aes(chain_first_session,id,alpha=chain_last_session-chain_first_session,color=dev_area))+geom_point()+facet_wrap(~dev_area)

ggplot(df_chains[!df_chains$dev_area_char %in% c('A','B')],
       aes(chain_first_session,weight=chain_last_session-chain_first_session,fill=dev_area))+
  geom_density(aes(y = after_stat(count)), adjust = 1)+ #weighted by occurances
  labs(x='session number',title = 'density of chains across sessions weighted by chain length')+
  facet_wrap(~dev_area)

```

## chain analysis {#chain-analysis}

- main predictor: chain_length
- main outcome:
  - help ("Ausmaß Hilfestellung (0-2)")
  - achieve ("Zielebene gemeistert (0-2)")
  - mastery ("aim expert level (0-2)")
- only consider chains not disrupted by lockdown; complete before or after

```{r chain analysis data preparation}

#describe outcome variables
psych::describe(df_chains$chain_length)
psych::describe(df_chains$chain_help_mean)
psych::describe(df_chains$chain_achieve_mean)
psych::describe(df_chains$chain_mastery_mean)

#non-normal distribution
hist(df_chains$chain_length)
hist(df_chains$chain_help_mean)
hist(df_chains$chain_achieve_mean)
hist(df_chains$chain_mastery_mean)

#prepare data 
lmm_data<-df_chains[df_chains$chain_after_lockdown_percentage %in% c(0,1),]

# function to plot model as APA table
fun_table_lmm<-function(lmm,caption){

lmm_summary<-data.frame(round(summary(lmm)$coefficients,3))
names(lmm_summary)<-c('estimate','SE','df','t-value','p-value')
variable<-c('intercept','chain length','full IQ','gender','age','after lockdown')
lmm_summary<-data.frame(variable,lmm_summary)

lmm_summary %>%
  kbl(caption = caption,row.names = F) %>%
  kable_classic(full_width = F, html_font = "Cambria")

}

```

### linear mixed models

```{r chain analysis frequentist}

#core models
lmm<-lmer(scale(chain_help_mean)~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
            as.factor(chain_after_lockdown_percentage)+(1|id),data=lmm_data)
fun_table_lmm(lmm,caption="help outcome: effect of chain length")

plot(lmm)
qqnorm(resid(lmm))
qqline(resid(lmm))

lmm<-lmer(scale(chain_achieve_mean)~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
            as.factor(chain_after_lockdown_percentage)+(1|id),data=lmm_data)
fun_table_lmm(lmm,caption="achieve outcome: effect of chain length")

plot(lmm)
qqnorm(resid(lmm))
qqline(resid(lmm))


lmm<-lmer(scale(chain_mastery_mean)~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
            as.factor(chain_after_lockdown_percentage)+(1|id),data=lmm_data)
fun_table_lmm(lmm,caption="mastery outcome: effect of chain length")

plot(lmm)
qqnorm(resid(lmm))
qqline(resid(lmm))


```

### linear mixed models - common dev areas {chain-analysis-restricted}

- only extract the most common developmental areas
- effects are confirmed when data is restricted to three most common dev areas

```{r chain analysis frequentist in common dev areas}

lmm_data_res<-lmm_data[lmm_data$dev_area %in% c('communication & speech (10.2)',
                                                'basic abilities (9)',
                                                'interaction & play (10.3)'),]

lmm<-lmer(scale(chain_help_mean)~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
            as.factor(chain_after_lockdown_percentage)+(1|id),data=lmm_data_res)
fun_table_lmm(lmm,caption="help outcome: effect of chain length")


lmm<-lmer(scale(chain_achieve_mean)~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
            as.factor(chain_after_lockdown_percentage)+(1|id),data=lmm_data_res)
fun_table_lmm(lmm,caption="achieve outcome: effect of chain length")

lmm<-lmer(scale(chain_mastery_mean)~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
            as.factor(chain_after_lockdown_percentage)+(1|id),data=lmm_data_res)
fun_table_lmm(lmm,caption="mastery outcome: effect of chain length")

```

### Bayesian ordinal mixed model

- assumes ordinal data
- consider the last session in a chain and its rating

```{r chain analysis Bayesian}

lmm_data$chain_help_last<-as.factor(lmm_data$chain_help_last)
lmm_data$chain_help_last<-ordered(lmm_data$chain_help_last, c(0,1,2))

# fit<-brm(chain_help_last~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
#             as.factor(chain_after_lockdown_percentage)+(1|id),
#          data=lmm_data,
#          family = cumulative('logit'),
#          chains = 4,
#          cores = 4
#          )
# 
# summary(fit)
# saveRDS(fit, file = paste0(getwd(),"/session data/data/brms_model_chain_help.rds"))


fit <- readRDS(paste0(root_path,"/session data/data/brms_model_chain_help.rds"))
summary(fit)
print('a 1 SD increase in chain length is associated with about a 9% increase in odds of being in less help category:')
exp(0.09)


# lmm_data$chain_achieve_last<-ordered(lmm_data$chain_achieve_last, c(0,1,2))
# fit2<-brm(chain_achieve_last~scale(chain_length)+scale(nonverbal_iq)+gender+scale(agedemo)+
#             as.factor(chain_after_lockdown_percentage)+(1|id),
#          data=lmm_data,
#          family = cumulative('logit'),
#          chains = 4,
#          cores = 4
#          )
# 
# summary(fit2)
# saveRDS(fit2, file = paste0(getwd(),"/session data/data/brms_model_chain_achieve.rds"))

fit2 <- readRDS(paste0(root_path,"/session data/data/brms_model_chain_achieve.rds"))
summary(fit2)
print('a 1 SD increase in chain length is associated with about a 18% increase in odds of being in higher achieve category:')
exp(0.17)


####need to compute chain_mastery_last

```



- clustern based on clusters and quality of clusters
- Prüfung der weiteren Angaben der Therapieziel im Verlauf der Therapie -> Stabilität bzw. Wechsel in andere Kategorie.


